
# Dependencies

```{r}

library(tidyverse)

source("functions.R")

```


```{r}

# results <- 
#   expand_grid(min = 1,
#               max = 7,
#               mean = seq(from = 1, to = 7, by = 1/10^2),
#               n = 30) |>
#   mutate(sd_min = pmap(list(mean, n), min_sd)) |>
#   unnest(sd_min) 
# # |>
# #   distinct(sd_min, .keep_all = TRUE)
# 
# ggplot(results, aes(mean, sd_min)) +
#   geom_line()

```

# Lookup tables

## Possible values

```{r}

combinations <- values_of_possible_means_and_sds(n = 10, min = 1, max = 7)
combinations

sprite_plot(n = 10, min = 1, max = 7)


values_plot(n = 10, min = 1, max = 5)
values_plot(n = 10, min = 1, max = 6)
values_plot(n = 10, min = 1, max = 7)

values_plot(n = 10, min = 1, max = 7)
values_plot(n = 15, min = 1, max = 7)
values_plot(n = 20, min = 1, max = 7)

```

## Limits

\TODO min_sd() function has always returned weird looking results, and this confirms that it produces different results to the limits function. Revisit min_sd().

```{r}

data_limits <- limits_of_possible_means_and_sds(n = 40, 
                                                min_score = 1, 
                                                max_score = 7)

```


```{r}

plot_values <- values_plot(n = 30, 
                           min_score = 1, 
                           max_score = 7)

temp <- limits_of_possible_means_and_sds(n = 30, 
                                         min_score = 1, 
                                         max_score = 7)

plot_values_limits <- values_plot(n = 30, min = 1, max = 7, n_digits_reported = 3) +
  geom_line(data = temp, aes(mean, sd_min)) +
  geom_line(data = temp, aes(mean, sd_max))

plot_limits <- limits_plot(n = 30, min_score = 1, max_score = 7, n_digits_reported = 3)

```

```{r}
plot_values
plot_values_limits
plot_limits
```


```{r}

limits_plot(n = 10, min_score = 1, max_score = 7, n_digits_reported = 3)
limits_plot(n = 20, min_score = 1, max_score = 7, n_digits_reported = 3)
limits_plot(n = 30, min_score = 1, max_score = 7, n_digits_reported = 3)
limits_plot(n = 40, min_score = 1, max_score = 7, n_digits_reported = 3)


limits_plot(n = 1000, min_score = 1, max_score = 7, n_digits_reported = 3)

```

# Boostrapping

## PHQ9

```{r}

data_phq9 <- read_csv("Flores-Kanter et al. - 2021 - GAD-7, PHQ-9 and PANAS-18.csv") |>
  select(starts_with("PHQ")) |>
  rownames_to_column(var = "id") 

data_phq9_meanscore <- data_phq9 |>
  pivot_longer(cols = starts_with("PHQ"),
               names_to = "item",
               values_to = "score") |>
  summarize(score = mean(score),
            .by = "id")

data_phq9_item9 <- data_phq9 |>
  select(id, score = PHQ9)

```

## PANAS

```{r}

data_panas <- read_csv("Flores-Kanter et al. - 2021 - GAD-7, PHQ-9 and PANAS-18.csv") |>
  select(starts_with("PANAS")) |>
  rownames_to_column(var = "id") 

data_panas_item9 <- data_panas |>
  select(id, score = PANAS9)

```

## BDI-II

```{r}

data_bdiii <- read_csv("squires et al - 2022 - BDI-II, BAI, Event-Related Rumination Inventory.csv") |>
  select(id = AnalysisID, starts_with("BDIII")) |>
  drop_na()

data_bdiii_meanscore <- data_bdiii |>
  pivot_longer(cols = starts_with("BDIII"),
               names_to = "item",
               values_to = "score") |>
  summarize(score = mean(score),
            .by = "id")

data_bdiii_item9 <- data_bdiii |>
  select(id, score = BDIII_9)


# run
result <- 
  bootstrap_summary_stats(data = data_bdiii_item9, 
                          sample_n = 25, 
                          iterations = 10000) |>
  dplyr::count(n, mean, sd, name = "frequency")


ggplot(result, aes(mean, sd, color = frequency)) +
  geom_point(shape = "square") +
  scale_color_viridis_c(begin = 0.2, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7)


# run
result_n10 <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  mutate(mean = round(mean, 2),
         sd = round(sd, 2)) |>
  dplyr::count(n, mean, sd, name = "frequency")




result2 <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 100, 
                          iterations = 10000) |>
  mutate(mean = round(mean, 2),
         sd = round(sd, 2)) |>
  dplyr::count(n, mean, sd, name = "frequency")





result3 <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 500, 
                          iterations = 10000) |>
  mutate(mean = round(mean, 2),
         sd = round(sd, 2)) |>
  dplyr::count(n, mean, sd, name = "frequency")


ggplot(result_n10, aes(mean, sd, color = frequency)) +
  geom_point(shape = "square") +
  scale_color_viridis_c(begin = 0.2, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7)

ggplot(result2, aes(mean, sd, color = frequency)) +
  geom_point(shape = "square") +
  scale_color_viridis_c(begin = 0.2, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7)

ggplot(result3, aes(mean, sd, color = frequency)) +
  geom_point(shape = "square") +
  scale_color_viridis_c(begin = 0.2, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7)

```

# combined bootstrapping and possible values

## BDI-II item 9

### Native

```{r}

possible_values <- 
  all_possible_means_and_sds(n = 10, 
                             min = 0, 
                             max = 3)

bootstrapped_values <- 
  bootstrap_summary_stats(data = data_bdiii_item9, 
                          sample_n = 10, 
                          iterations = 10000) |>
  dplyr::count(n, mean, sd, name = "frequency")

ggplot() +
  geom_point(data = possible_values, aes(mean, sd), shape = "square", color = "grey") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7) +
  theme_linedraw() +
  ggtitle("SPRITE plot (Heathers et al., 2018)")

ggplot() +
  geom_point(data = possible_values, aes(mean, sd), shape = "square", color = "grey") +
  geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7) +
  theme_linedraw() +
  ggtitle("TIDES (Hussey et al., 2024)")

```

### Standardized

```{r}

possible_values_std <- possible_values |>
  mutate(mean = (mean - min(mean)) / (max(mean) - min(mean)),
         sd = sd / max(sd))

bootstrapped_values_std <- bootstrapped_values |>
  mutate(mean = (mean - min(possible_values$mean)) / (max(possible_values$mean) - min(possible_values$mean)),
         sd = sd / max(possible_values$sd))

ggplot() +
  geom_point(data = possible_values_std, aes(mean, sd/max(sd)), shape = "square", color = "grey") +
  geom_point(data = bootstrapped_values_std, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlab("Standardized mean") +
  ylab("Standardized SD") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_linedraw()

```

## PANAS item 9

### Native

```{r}

possible_values_panas9 <- 
  all_possible_means_and_sds(n = 10, 
                             min = 1, 
                             max = 5)

bootstrapped_values_panas9 <- 
  bootstrap_summary_stats(data = data_panas_item9, 
                          sample_n = 10, 
                          iterations = 10000) |>
  dplyr::count(n, mean, sd, name = "frequency")

ggplot() +
  geom_point(data = possible_values_panas9, aes(mean, sd), shape = "square", color = "grey") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlim(1, 5) +
  ylim(0, 2.2) +
  theme_linedraw() +
  ggtitle("SPRITE plot (Heathers et al., 2018)")

ggplot() +
  geom_point(data = possible_values_panas9, aes(mean, sd), shape = "square", color = "grey") +
  geom_point(data = bootstrapped_values_panas9, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlim(1, 5) +
  ylim(0, 2.2) +
  theme_linedraw() +
  ggtitle("TIDES (Hussey et al., 2024)")

```

### Standardized

```{r}

possible_values_std_panas9 <- possible_values_panas9 |>
  mutate(mean = (mean - min(mean)) / (max(mean) - min(mean)),
         sd = sd / max(sd))

bootstrapped_values_std_panas9 <- bootstrapped_values_panas9 |>
  mutate(mean = (mean - min(possible_values_panas9$mean)) / (max(possible_values_panas9$mean) - min(possible_values_panas9$mean)),
         sd = sd / max(possible_values_panas9$sd))

ggplot() +
  geom_point(data = possible_values_std_panas9, aes(mean, sd/max(sd)), shape = "square", color = "grey") +
  geom_point(data = bootstrapped_values_std_panas9, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlab("Standardized mean") +
  ylab("Standardized SD") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_linedraw() +
  ggtitle("TIDES (Hussey et al., 2024)\nBDI-II item 9")

ggplot() +
  geom_point(data = possible_values_std, aes(mean, sd/max(sd)), shape = "square", color = "grey") +
  geom_point(data = bootstrapped_values_std, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlab("Standardized mean") +
  ylab("Standardized SD") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_linedraw() +
  ggtitle("TIDES (Hussey et al., 2024)\nPANAS item 9")

```

# Bootstrapped full scale

## PHQ-9 full scale

### N=10

#### Native

```{r}

bootstrapped_values_phq9 <- 
  bootstrap_summary_stats(data = data_phq9_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  mutate(mean = janitor::round_half_up(mean*9, 3),
         sd = janitor::round_half_up(sd*9, 3))

# ggplot() +
#   geom_point(data = bootstrapped_values_phq9, aes(mean, sd, color = frequency), shape = "square") +
#   scale_color_viridis_c(begin = 0.5, end = 1) +
#   xlim(0, 3) +
#   ylim(0, 1.7) +
#   theme_linedraw() +
#   ggtitle("TIDES (Hussey et al., 2024)")

limits_plot(n = 10, min_score = 1*9, max_score = 5*9, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_phq9, aes(mean, sd, color = frequency), shape = "square")



bootstrapped_values_phq9_rounded <- 
  bootstrap_summary_stats(data = data_phq9_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  mutate(mean = janitor::round_half_up(mean*9, 1),
         sd = janitor::round_half_up(sd*9, 1))

limits_plot(n = 10, min_score = 1*9, max_score = 5*9, n_digits_reported = 1) +
  geom_point(data = bootstrapped_values_phq9_rounded, aes(mean, sd, color = frequency), shape = "square")

```

#### Standardized

\TODO there is a manual bodge to take the multiple items into account - needs fixing

\TODO think about how the rounding is done inside bootstrap_summary_stats()

```{r}

bootstrapped_values_phq9 <- 
  bootstrap_summary_stats(data = data_phq9_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  # manual bodge
  mutate(mean = mean*9,
         sd = sd*9)

bootstrapped_values_phq9_std <- range_normalize(data = bootstrapped_values_phq9,
                                                n = 10,
                                                min_score = 1*9,
                                                max_score = 5*9)

limits_plot(n = 10, min_score = 1*9, max_score = 5*9, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_phq9, aes(mean, sd, color = frequency), shape = "square") 

standardized_limits_plot(n = 10, min_score = 1*9, max_score = 5*9, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_phq9_std, aes(mean, sd, color = frequency), shape = "square") 

```

### N=100

#### Native

```{r}

# bootstrapped_values <- 
#   bootstrap_summary_stats(data = data_phq9_meanscore, 
#                           sample_n = 100, 
#                           iterations = 10000) |>
#   dplyr::count(n, mean, sd, name = "frequency")

# ggplot() +
#   geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") +
#   scale_color_viridis_c(begin = 0.5, end = 1) +
#   xlim(0, 3) +
#   ylim(0, 1.7) +
#   theme_linedraw() +
#   ggtitle("TIDES (Hussey et al., 2024)")

```

#### Standardized

```{r}

bootstrapped_values_phq9 <- 
  bootstrap_summary_stats(data = data_phq9_meanscore, 
                          sample_n = 100, 
                          iterations = 10000) |>
  # manual bodge
  mutate(mean = mean*9,
         sd = sd*9)

bootstrapped_values_phq9_std <- range_normalize(data = bootstrapped_values_phq9,
                                                n = 100,
                                                min_score = 1*9,
                                                max_score = 5*9)

limits_plot(n = 100, min_score = 1*9, max_score = 5*9, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_phq9, aes(mean, sd, color = frequency), shape = "square") 

standardized_limits_plot(n = 100, min_score = 1*9, max_score = 5*9, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_phq9_std, aes(mean, sd, color = frequency), shape = "square") 

```

## BDI-II full scale

### N=10

#### Native

```{r}

bootstrapped_values <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  mutate(mean = janitor::round_half_up(mean*21, 3),
         sd = janitor::round_half_up(sd*21, 3))

# ggplot() +
#   geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") +
#   scale_color_viridis_c(begin = 0.5, end = 1) +
#   xlim(0, 3) +
#   ylim(0, 1.7) +
#   theme_linedraw() +
#   ggtitle("TIDES (Hussey et al., 2024)")

limits_plot(n = 10, min_score = 0, max_score = 3*21, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square")



bootstrapped_values_rounded <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  mutate(mean = janitor::round_half_up(mean*21, 1),
         sd = janitor::round_half_up(sd*21, 1))

limits_plot(n = 10, min_score = 0, max_score = 3*21, n_digits_reported = 1) +
  geom_point(data = bootstrapped_values_rounded, aes(mean, sd, color = frequency), shape = "square")

```

#### Standardized

\TODO there is a manual bodge to take the multiple items into account - needs fixing

\TODO think about how the rounding is done inside bootstrap_summary_stats()

```{r}

bootstrapped_values <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 10, 
                          iterations = 10000) |>
  # manual bodge
  mutate(mean = mean*21,
         sd = sd*21)

bootstrapped_values_std <- range_normalize(data = bootstrapped_values,
                                           n = 10,
                                           min_score = 0,
                                           max_score = 63)

limits_plot(n = 10, min_score = 0, max_score = 63, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") 

standardized_limits_plot(n = 10, min_score = 0, max_score = 63, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_std, aes(mean, sd, color = frequency), shape = "square") 

counts_pomp <- bootstrapped_values_std |>
  mutate(mean = janitor::round_half_up(mean, 2),
         sd = janitor::round_half_up(sd, 2)) 

counts_pomp |>
  distinct(mean) |>
  dplyr::count(name = "n_mean")

counts_pomp |>
  distinct(sd) |>
  dplyr::count(name = "n_sd")

counts_pomp |>
  distinct(mean, sd) |>
  dplyr::count(name = "n_mean_sd_pairs")

```

### N=100

#### Native

```{r}

bootstrapped_values <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 100, 
                          iterations = 10000) |>
  dplyr::count(n, mean, sd, name = "frequency")

ggplot() +
  geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") +
  scale_color_viridis_c(begin = 0.5, end = 1) +
  xlim(0, 3) +
  ylim(0, 1.7) +
  theme_linedraw() +
  ggtitle("TIDES (Hussey et al., 2024)")

```

#### Standardized

```{r}

bootstrapped_values <- 
  bootstrap_summary_stats(data = data_bdiii_meanscore, 
                          sample_n = 100, 
                          iterations = 10000,
                          n_digits_reported = 3) |>
  # manual bodge
  mutate(mean = mean*21,
         sd = sd*21)

bootstrapped_values_std <- range_normalize(data = bootstrapped_values,
                                           n = 100,
                                           min_score = 0,
                                           max_score = 63)

limits_plot(n = 100, min_score = 0, max_score = 63, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values, aes(mean, sd, color = frequency), shape = "square") 

standardized_limits_plot(n = 100, min_score = 0, max_score = 63, n_digits_reported = 3) +
  geom_point(data = bootstrapped_values_std, aes(mean, sd, color = frequency), shape = "square") 

counts_pomp <- bootstrapped_values_std |>
  mutate(mean = janitor::round_half_up(mean, 2),
         sd = janitor::round_half_up(sd, 2)) 

counts_pomp |>
  distinct(mean) |>
  dplyr::count(name = "n_mean")

counts_pomp |>
  distinct(sd) |>
  dplyr::count(name = "n_sd")

counts_pomp |>
  distinct(mean, sd) |>
  dplyr::count(name = "n_mean_sd_pairs")

```




