---
title: "SPRITE"
author: "Ian Hussey"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPRITE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

library(tidyverse)
library(janitor)
library(rsprite2)
library(knitr)
library(kableExtra)

min_decimals <- function(x, digits = 2) {
  sprintf(paste0("%.", digits, "f"), x)
}

```

# SPRITE

Note that data is hard coded into the RMarkdown file. It could also be loaded from disk, e.g., from a correctly formatted .csv file (with the headers label, mean, sd, n, and items) to a tibble via readr::read_csv(). 

Note that datasets are only generated using the first row of the summary_statistics dataframe. You can alter the code below to change which row is used.

```{r}

# # enter your data here. Here is some example data
# summary_statistics <- 
#   tibble(
#     label = "age",
#     mean  = 1,
#     sd    = 0,
#     n     = 10,
#     items = 1 # number of items in the scale used to measure this variable. For a single item measure like "what is your age?", items = 1. For a multi item scale, e.g., the BDI-II, this would be the number of items (e.g., 21).
#   ) |>
#   # these conversions to character and, below, back to numeric are done because the code is also common to the GRIM and GRIMMER tests, which require character data.
#   mutate(x = as.character(min_decimals(mean)),
#          sd = as.character(min_decimals(sd))) 
# 
# # summary statistics to generate data from
# # note that when the summary_statistics data frame has multiple rows, only the first is used. Alter the code here to change this behavior, e.g., from [1] to [2].
# sprite_parameters <- 
#   set_parameters(mean = summary_statistics$mean[1], 
#                  m_prec = 2, # number of decimal place M was reported to
#                  sd = as.numeric(summary_statistics$sd[1]), 
#                  sd_prec = 2, # number of decimal place SD was reported to
#                  n_obs = summary_statistics$n[1],
#                  min_val = 1, # lowest possible value of variable, if applicable. Eg. age could not be less than 18
#                  max_val = 7)  # highest possible value of variable, if applicable. Eg. age unlikely to be above 120
# 
# # generate data
# sprite_results <- 
#   find_possible_distributions(sprite_parameters, 
#                               seed = 42, # set seed for random number generator for reproducibility
#                               n_distributions = 100)

```

```{r}

# summary statistics to generate data from
# note that when the summary_statistics data frame has multiple rows, only the first is used. Alter the code here to change this behavior, e.g., from [1] to [2].
sprite_parameters <- 
  set_parameters(mean = 1.00, 
                 m_prec = 2, # number of decimal place M was reported to
                 sd = 0.00, 
                 sd_prec = 2, # number of decimal place SD was reported to
                 n_obs = 10,
                 min_val = 1, # lowest possible value of variable, if applicable. Eg. age could not be less than 18
                 max_val = 7)  # highest possible value of variable, if applicable. Eg. age unlikely to be above 120

# generate data
sprite_results <- 
  find_possible_distributions(sprite_parameters, 
                              seed = 42, # set seed for random number generator for reproducibility
                              n_distributions = 1000)

```

```{r}

# error in .sd_limits: this should return TRUE
GRIMMER_test(mean = 7, sd = 0, n_obs = 100, m_prec = 2, sd_prec = 2, n_items = 1, min_val = 1, max_val = 7)

```


```{r}

library(purrr)

safe_GRIMMER_test = possibly(GRIMMER_test, otherwise = FALSE)

dat <- 
  # expand_grid(mean = seq(from = 1, to = 3, by = 0.01),
  #             sd = seq(from = 0, to = 1.5, by = 0.01)) |>
  expand_grid(mean = seq(from = 1, to = 7, by = 0.01),
              sd = seq(from = 0, to = 3.5, by = 0.01)) |>
  mutate(n_obs   = 14,
         m_prec  = 2,
         sd_prec = 2,
         n_items = 1,
         # min_val = min(mean), 
         # max_val = max(mean)) |>
         min_val = 1, 
         max_val = 7) |>
  mutate(grim    = pmap(list(mean, n_obs, m_prec, n_items), GRIM_test)) |>
  mutate(grimmer = pmap(list(mean, sd, n_obs, m_prec, sd_prec, n_items, min_val, max_val), safe_GRIMMER_test)) |>
  unnest(grim) |>
  unnest(grimmer)
  #filter(grim & grimmer)

```

# Umbrella plot

```{r fig.height=7, fig.width=10}

dat |>
  filter(grim) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

dat |>
  filter(grimmer) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

dat |>
  filter(grim & grimmer) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

```

# ??

```{r}

sprite_results <-
  find_possible_distributions(set_parameters(mean    = 1.1,
                                             sd      = 0.32,
                                             n_obs   = 10,
                                             min_val = 1,
                                             max_val = 3,
                                             m_prec  = 2,
                                             sd_prec = 2),
                              seed = 42,
                              n_distributions = 10)

sprite <- function(mean, sd, n_obs, min_val, max_val){
  find_possible_distributions(set_parameters(mean    = mean,
                                             sd      = sd,
                                             n_obs   = n_obs,
                                             min_val = min_val,
                                             max_val = max_val,
                                             m_prec  = 2,
                                             sd_prec = 2),
                              seed = 42,
                              n_distributions = 10) |>
    select(-mean, -sd)
}

dat2 <- dat |>
  mutate(sprite = pmap(list(mean, sd, n_obs, min_val, max_val), sprite)) |>
  unnest(sprite)

dat2 |>
  count(mean, sd) |>
  arrange(desc(n))

```

## Plot type 1

Useful to see the min and max scores that could generate the summary statistics. Harder to interpret multimodality.

```{r}

plot_distributions(sprite_results)

```

## Plot type 2

Useful to diagnose multimodality and other weird distributions, and min/max range, but can obscures the distribution of individual datasets. After all, only one of these datasets is true (if a real dataset did exist, and possibly not within the 100 generated as SPRITE isn't exhaustive).

```{r fig.height=15, fig.width=6}

enframe(sprite_results$distribution) %>%
  unnest(cols = c(value)) %>%
  group_by(name) %>%
  rowid_to_column() %>%
  ungroup() %>%
  ggplot(aes(value)) + 
  geom_histogram(aes(fill = "all", binwidth = 1, position = "identity", boundary = 0)) +
  scale_fill_manual(values = "#702963") +
  theme_linedraw() +
  facet_wrap(~ name, ncol = 6) +
  theme(legend.position = "none") +
  xlab("Response") +
  ggtitle("Some of the possible distributions\nthat conform to the reported N, M, and SD")

```

## Plot type 3

Useful to see individual datasets and their distributions. E.g., if there is good reason to believe the real data should be right skewed (e.g., it is BDI-II data), inspection of individual datasets can help you assess whether this is plausible given these summary statistics.

```{r}

enframe(sprite_results$distribution) %>%
  unnest(cols = c(value)) %>%
  group_by(name) %>%
  rowid_to_column() %>%
  ungroup() %>%
  ggplot(aes(value)) + 
  geom_histogram(aes(group = as.factor(name)), alpha = 0.1, fill = "#702963", binwidth = 1, position = "identity", boundary = 0) +
  theme_linedraw() +
  xlab("Response") +
  ggtitle("Some of the possible distributions\nthat conform to the reported N, M, and SD")

```


