---
title: "Certifying integer possibility"
subtitle: "CLOSURE reconstruction and analytic certification compared"
author: "Ian Hussey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Certifying integer possibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_unsum <- requireNamespace("unsum", quietly = TRUE)
```

```{r message=FALSE}
library(strait)
```

## Necessary is not sufficient

`brimmer()` screens a reported (mean, SD, *n*) triple against the closed-form
SD bounds, and defers GRIM and GRIMMER to `scrutiny`. Each of those tests is
*necessary*: failing one proves the report impossible. Passing all of them does
not prove it possible.

A small residual set of tuples clears every closed-form test and still has no
integer solution. They cluster near the mean-conditional ceiling, where the
bounds are loosest relative to the discrete lattice underneath them. On a 1--5
scale at *n* = 9, reporting to one decimal place:

```{r}
um <- umbrella_data(n = 9, l = 1, u = 5, digits = 1, Z = "integer")
screened <- sum(um$consistent)

cert <- certify(l = 1, u = 5, n = 9, mean = um$mean, sd = um$sd, digits = 1)

c(cells             = nrow(um),
  passes_screen     = screened,
  certified         = sum(cert$possible),
  blind_spot        = sum(um$consistent & !cert$possible),
  wrongly_screened  = sum(!um$consistent & cert$possible))
```

The last number is the one that matters for trusting the screen: it is zero.
The closed-form tests never reject a report that real integer data can produce.
The `blind_spot` count is the price of that safety — reports the screen lets
through which are nonetheless impossible.

## Two ways to settle them

**CLOSURE**, in the `unsum` package, reconstructs datasets. Given a report it
searches for integer samples matching it, and returns them. If the search
completes with nothing found, the report is impossible.

**`certify()`** answers the same question analytically. It enumerates the exact
attainable `(mean, sd)` lattice for the design by dynamic programming over the
reachable *(sum, sum of squares)* states, rounds that lattice to the reporting
precision, and tests membership. No dataset is ever constructed — the lattice
records *which* pairs are reachable, not *how*.

Both are exact. They differ in what they return and in what their cost depends
on.

```{r}
# a report a real 1-5 sample can produce
certify(l = 1, u = 5, n = 9, mean = 3.0, sd = 1.0, digits = 1)

# a blind-spot cell: passes the bounds, GRIM and GRIMMER, but no sample exists
certify(l = 1, u = 5, n = 9, mean = 1.3, sd = 0.9, digits = 1)
```

`possible` is the verdict; `rules` names the rounding conventions under which
the report is reachable.

Because one lattice is built per design and reused, certifying a whole table of
reports costs barely more than certifying one:

```{r}
certify(l = 1, u = 5, n = 9, digits = 1,
        mean = c(3.0, 1.3, 2.5, 2.0),
        sd   = c(1.0, 0.9, 1.2, 1.1))
```

## They do not disagree

Run over every cell of the umbrella above, the two methods return identical
verdicts:

```{r eval=has_unsum}
closure_says <- vapply(seq_len(nrow(um)), function(i) {
  r <- try(suppressWarnings(unsum::closure_generate(
    mean = sprintf("%.1f", um$mean[i]), sd = sprintf("%.1f", um$sd[i]),
    n = 9, scale_min = 1, scale_max = 5, rounding = "up_or_down")),
    silent = TRUE)
  if (inherits(r, "try-error")) FALSE else nrow(r$results) > 0
}, logical(1))

c(cells = nrow(um),
  agree = sum(closure_says == cert$possible),
  disagree = sum(closure_says != cert$possible))
```

Across six designs spanning 1--5 and 1--7 scales and *n* from 9 to 20 — about
5,700 cells in total — there were no disagreements at all.

## Where the costs go

CLOSURE's work is proportional to the number of datasets it enumerates, which
grows combinatorially with *n* and with the width of the scale. `certify()`'s
work depends only on `l`, `u` and *n*, and is independent of how many datasets
happen to exist; the whole grid is certified in one pass.

Measured on the full umbrella grid of each design (macOS, R 4.5.2):

| *n* | scale | cells | `certify()` | CLOSURE | speed-up |
|----:|:------|------:|------------:|--------:|---------:|
|   9 | 1--5  |   610 |     0.003 s |  1.80 s |    ~700x |
|  12 | 1--5  |   684 |     0.004 s |  1.72 s |    ~430x |
|  15 | 1--5  |   693 |     0.002 s |  1.62 s |    ~840x |
|  20 | 1--5  |   706 |     0.002 s |  1.72 s |    ~700x |
|  12 | 1--7  |  1497 |     0.004 s |  4.12 s |   ~1170x |
|  15 | 1--7  |  1499 |     0.007 s |  4.96 s |    ~740x |

The lattice itself scales well past the point where reconstruction becomes
uncomfortable. On a 1--7 scale it reaches *n* = 400 — 2.8 million distinct
attainable tuples — in about 23 seconds, and a 1--5 scale at *n* = 200 takes
under a second. A `max_cells` guard (default `2e7`) makes the enumeration error
rather than exhaust memory when a design is genuinely too large.

This might reasonably be called an **analytic CLOSURE certification**: the same
certificate CLOSURE provides, obtained by construction of the reachable set
rather than by search of the sample space.

## Choosing between them

Use `certify()` when the verdict is the deliverable — screening a literature,
triaging a table of reported statistics, or settling whether a specific report
is arithmetically possible.

Use CLOSURE when you need the *witnesses*: the actual candidate datasets, their
distributional shape, or the count of solutions. `certify()` cannot supply
those, by construction. The two compose naturally — `brimmer()` as a fast first
screen, `certify()` to settle what survives it, and CLOSURE when a surviving
case needs its datasets examined.

## Rounding, and the direction of the proof

The two directions of the certificate are not symmetric.

A **hit** proves possibility outright: some sample rounds to exactly this
report, under the rule that produced the hit.

A **miss** proves impossibility only *relative to the rounding rules admitted*.
A report unreachable under one convention may be reachable under another, so
the default admits both `"half_up"` and `"half_down"`, spanning the usual
ambiguity about how a source rounded its halves:

```{r}
# c(rep(0, 9), 1, 1, 1) is a real sample of 12 integers on 0-6:
# mean 0.25, sd 0.4523, reportable as 0.2 and 0.5
x <- c(rep(0, 9), 1, 1, 1)
c(mean = mean(x), sd = sd(x))

certify(l = 0, u = 6, n = 12, mean = 0.2, sd = 0.5, digits = 1)
```

The `rules` column shows this report is reachable only when halves round down —
0.25 becomes "0.2" under `half_down` but "0.3" under `half_up`. Restricting
`rounding` to a convention the source did not use would report this real sample
as impossible. Narrow the rule set only when the source's convention is
actually known; for a forensic claim, a false impossibility is the costly
error.
