---
title: "strait"
subtitle: "Bounds tests for reported summary statistics"
author: "Ian Hussey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{strait}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
options(scipen = 999)
```

```{r}
library(strait)
```

## The idea

When a measure has a known minimum and maximum, a reported mean constrains the
sample standard deviation that is arithmetically possible: the two summary
statistics are *not* independent. `strait` computes, in closed form, the smallest
and largest sample SD consistent with a chosen set of constraints, and flags
reported values that fall outside them.

The constraints are nested — supplying more of them can only narrow the bounds.

```{r}
sd_bounds(l = 1, u = 5)                          # scale limits only
sd_bounds(l = 1, u = 5, n = 9)                   # + sample size (parity ceiling)
sd_bounds(l = 1, u = 5, n = 9, mean = 2)         # + mean (sharp ceiling)
```

The `min_rule` / `max_rule` columns name which bound binds.

## Granularity

If the data are integers (e.g. single Likert items), the SD also has a nonzero
*floor* — data cannot be perfectly smooth. `granularity = "integer"` requires every
observation to be a whole number (and the mean to be GRIM-consistent);
`granularity = "quasiinteger"` relaxes this to "all but one", which is defined at every
mean and so never has GRIM holes.

```{r}
sd_bounds(l = 1, u = 5, n = 9, mean = 2.44, granularity = "quasiinteger")
```

## Checking a reported statistic

Three tests, named for how much they ask, and nested so that each contains the
one before it:

| test | asks | verdict |
|---|---|---|
| `brim()` | is the reported mean attainable? | necessary only |
| `brimmer()` | and is the reported SD attainable with it? | necessary only |
| `brimmest()` | is the pair jointly attainable by real integer data? | necessary **and** sufficient |

`brim()` is the mean-side test on its own. It needs no SD, and asks only whether
the reported mean could have arisen at all: within the scale limits, within the
narrower band implied by any reported attained extremes, and — under
`granularity = "integer"` — GRIM-consistent.

```{r}
# an ordinary mean
brim(l = 1, u = 7, n = 30, mean = 2.97, digits_mean = 2, granularity = "integer")

# GRIM-impossible: 30 x 3.51 is not a whole number
brim(l = 1, u = 7, n = 30, mean = 3.51, digits_mean = 2, granularity = "integer")
```

The failures are named separately, so an out-of-range mean (`in_scale_range`)
is never confused with a granularity-impossible one (`grim`). `band_lo` and
`band_hi` report the feasible band the mean was tested against; reporting an
attained minimum and maximum narrows it.

```{r}
brim(a = 1, b = 7, n = 30, mean = 1.10, digits_mean = 2)[
  , c("consistent", "failed_tests", "band_lo", "band_hi")]
```

`brimmer()` adds the SD. Give it the reported values as they appear in a paper —
numbers plus their decimal places — and it unrounds them, applies the bounds
test, and (for integer data) defers the GRIM and GRIMMER verdicts to the
`scrutiny` package.

```{r}
# an ordinary report on a 1-7 scale
brimmer(l = 1, u = 7, n = 30, mean = 2.97, digits_mean = 2,
        sd = 2.83, digits_sd = 2, granularity = "integer")

# an SD too small to be possible: caught by the bounds test
brimmer(l = 1, u = 7, n = 30, mean = 2.97, digits_mean = 2,
        sd = 0.10, digits_sd = 2, granularity = "quasiinteger")
```

Each of these tests is *necessary*: failing one proves the report impossible. None
is sufficient, so a pass means "not ruled out" rather than "achievable" — a small
residue of reports clears every test and still corresponds to no integer data
set. `brimmest()` settles those; see [Settling what survives].

The check also returns percent-of-maximum-possible (POMP) transforms:
`pomp_mean`, a linear `pomp_sd_parity` (SD as a fraction of the mean-agnostic
maximum), and `pomp_sd_sharp` (the SD's position within the sharp
mean-conditional band: 0 at the floor, 1 at the ceiling, outside `[0, 1]` when
impossible).

## Across a data frame

`brimmer_map()` applies the check to each row of a data frame.
Columns named like the arguments are read per row; anything passed as a constant
is broadcast. Repeated constraint tuples are computed once and reused.

```{r}
reports <- tibble::tibble(
  mean = c(2.97, 4.50, 3.10),
  sd   = c(2.83, 0.10, 1.90)
)
brimmer_map(reports, l = 1, u = 7, n = 30,
                 digits_mean = 2, digits_sd = 2, granularity = "integer")[
  , c("mean", "sd", "consistent", "failed_tests")]
```

## Visualising the feasible envelope

`sd_bounds_curve()` traces the floor and ceiling across the mean;
`plot_sd_bounds()` draws the band and, optionally, reported points coloured by
consistency.

```{r}
curve  <- sd_bounds_curve(l = 1, u = 7, n = 30, granularity = "quasiinteger")
points <- brimmer_map(
  tibble::tibble(mean = c(3.0, 4.0, 2.5), sd = c(2.5, 0.2, 3.6)),
  l = 1, u = 7, n = 30, digits_mean = 1, digits_sd = 1, granularity = "quasiinteger")

plot_sd_bounds(curve, points = points)
```

Shading marks what is *ruled out*, so the clear region is the answer to the
question the figure was asked. `shade = "inside"` fills the feasible band
instead, and `expand` sets the margin as a proportion of the scale width `u - l`
rather than a fixed number of SD units, so a 1–7 scale and a 0–100 one are
padded alike.

```{r}
plot_sd_bounds(curve, points = points, shade = "inside", expand = 0.06)
```

On the POMP scale, `reference = "parity"` keeps the geometry undistorted (a
linear rescaling), while `reference = "sharp"` maps the feasible region to the
unit square so a point's height is its position within the band.

```{r}
plot_sd_bounds_pomp(curve, points = points, reference = "sharp")
```

## The nested constraints, drawn

Each constraint set in the framework is a `rule` that `plot_sd_region()` can
draw on its own, which is the quickest way to see what adding a constraint buys.

```{r}
plot_sd_region(l = 1, u = 5, n = 7, rule = "muilwijk",
               title = "Muilwijk's ceiling, as originally stated")
plot_sd_region(l = 1, u = 5, n = 7, rule = "quasi",
               title = "Both bounds sharp, and GRIM-free")
```

`"muilwijk"` (an alias for `"mean"`) lets the counts at each limit be
fractional. They cannot be, and `sd_delta()` is the price of rounding them: the
fraction of that ceiling actually reachable. It is 1 exactly when both counts
come out whole, and falls toward 0 near a limit, where the ceiling is least
attainable.

```{r}
sd_delta(mean = c(3, 2, 1.07), n = 7, lower = 1, upper = 5)

# which is exactly the gap between the two ceilings the package exports
sd_max_structure_s(3, 7, 1, 5)^2
sd_max_muilwijk(3, 7, 1, 5)^2 * sd_delta(3, 7, 1, 5)
```

Some rules leave stretches with no feasible SD at all — with a reported alpha
there are means near each limit at which no composite exists. `band_polygon()`
turns a band into closed rings so those stretches stay shaded rather than
becoming blank, which would wrongly suggest any SD were possible there.

```{r}
d <- sd_region_data(0, 3, 7, rule = "alpha", n_items = 2, alpha = 0.70)
c(means = nrow(d), undefined = sum(is.na(d$lo)),
  rings = length(unique(band_polygon(d, by = 3 / 1000)$ring)))

plot_sd_region(0, 3, 7, rule = "alpha", n_items = 2, alpha = 0.70,
               title = "A reported alpha: three rings, not one band")
```

## The umbrella

`umbrella_data()` builds the full grid of reported (mean, SD) pairs at a given
reporting precision, tagging each as consistent, GRIMMER-inconsistent, or out of
bounds. `plot_umbrella()` draws only the consistent ones by default, since every
other cell is a value that cannot be reported.

```{r}
um  <- umbrella_data(n = 14, l = 1, u = 7, digits = 2)
cur <- sd_bounds_curve(l = 1, u = 7, n = 14, granularity = "quasiinteger")
plot_umbrella(um, curve = cur)
```

The vertical striping is the real structure: the means an integer sum can round
to. `style = "tiles"` instead draws every cell of the grid, separating what the
bounds rule out from what GRIMMER additionally rules out — useful for exposition,
though it spends most of its ink on impossible values.

```{r}
plot_umbrella(um, curve = cur, style = "tiles")
```

## Settling what survives

The umbrella above is drawn from the *screen*, so some of the cells it marks
consistent are nonetheless unreachable. `brimmest()` decides the question
outright: it enumerates the exact attainable `(mean, SD)` lattice for the design
by dynamic programming, rounds it to the reporting precision, and tests
membership. No data set is constructed — the lattice records which pairs are
reachable, not how.

```{r}
# a report a real 1-5 sample can produce
brimmest(l = 1, u = 5, n = 9, mean = 3.0, sd = 1.0, digits_mean = 1, digits_sd = 1)

# a blind-spot cell: it passes the bounds, GRIM and GRIMMER, yet no sample exists
brimmest(l = 1, u = 5, n = 9, mean = 1.3, sd = 0.9, digits_mean = 1, digits_sd = 1)
```

`possible` is the verdict and `rules` names the rounding conventions under which
the report is reachable. One lattice is built per design and reused, so
certifying a whole table costs barely more than certifying one report.

```{r}
brimmest(l = 1, u = 5, n = 9, digits_mean = 1, digits_sd = 1,
        mean = c(3.0, 1.3, 2.5, 2.0),
        sd   = c(1.0, 0.9, 1.2, 1.1))
```

`brimmest_map()` is the data frame form, mirroring `brimmer_map()`:
columns named like the arguments are read per row, constants are broadcast.
Rows are grouped by design and each group certified in one call, so the reuse
above still applies when the design itself varies down the table.

```{r}
mixed <- tibble::tibble(mean = c(3.0, 1.3, 4.0, 3.0),
                    sd   = c(1.0, 0.9, 1.5, 1.0),
                    n    = c(9, 9, 12, 9))
brimmest_map(mixed, l = 1, u = 5, digits_mean = 1, digits_sd = 1)
```

The two directions of the certificate are not symmetric. A *hit* proves
possibility outright: some sample rounds to exactly this report. A *miss* proves
impossibility only relative to the rounding rules admitted, which is why the
default admits both `"half_up"` and `"half_down"`. Narrowing `rounding` to a
convention the source did not use risks calling a real sample impossible, and in
forensic work that is the costly error.

`brimmest()` returns a verdict, not witnesses. When the actual candidate data sets
matter — their shape, or how many there are — reconstruct them with
[unsum](https://lhdjung.github.io/unsum/), whose CLOSURE algorithm `brimmest()`
agrees with cell for cell.

## Multi-item scales and composites

`scoring` and `n_items` describe how response items form the score. Sum scores
are integer-gridded (`scoring = "sumscored"`); item means live on a finer
`1/n_items` grid (`scoring = "meanscored"`). A reported Cronbach's alpha
tightens the bounds of a composite further.

```{r}
# a 3-item composite reported as an item mean, with alpha
sd_bounds(l = 1, u = 5, n = 20, mean = 3, granularity = "integer",
          scoring = "meanscored", n_items = 3)$max_sd

sd_bounds(l = 1, u = 5, n = 20, mean = 3, granularity = "integer",
          scoring = "meanscored", n_items = 3, alpha = 0.7)$max_sd
```
