---
title: "tides"
subtitle: "Bounds of the standard deviation for truncated data"
author: "Ian Hussey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
options(scipen = 999)
```

```{r}
library(tides)
```

## The idea

When a measure has a known minimum and maximum, a reported mean constrains the
sample standard deviation that is arithmetically possible: the two summary
statistics are *not* independent. `tides` computes, in closed form, the smallest
and largest sample SD consistent with a chosen set of constraints, and flags
reported values that fall outside them.

The constraints are nested — supplying more of them can only narrow the bounds.

```{r}
sd_bounds(l = 1, u = 5)                          # scale limits only
sd_bounds(l = 1, u = 5, n = 9)                   # + sample size (parity ceiling)
sd_bounds(l = 1, u = 5, n = 9, mean = 2)         # + mean (sharp ceiling)
```

The `min_rule` / `max_rule` columns name which bound binds.

## Granularity

If the data are integers (e.g. single Likert items), the SD also has a nonzero
*floor* — data cannot be perfectly smooth. `Z = "integer"` requires every
observation to be a whole number (and the mean to be GRIM-consistent);
`Z = "quasiinteger"` relaxes this to "all but one", which is defined at every
mean and so never has GRIM holes.

```{r}
sd_bounds(l = 1, u = 5, n = 9, mean = 2.44, Z = "quasiinteger")
```

## Checking a reported statistic

`sd_bounds_check()` turns the bounds into a consistency verdict. Give it the
reported values as they appear in a paper — numbers plus their decimal places —
and it unrounds them, applies the bounds test, and (for integer data) defers the
GRIM and GRIMMER verdicts to the `scrutiny` package.

```{r}
# an ordinary report on a 1-7 scale
sd_bounds_check(l = 1, u = 7, n = 30, mean = 2.97, mean_digits = 2,
                sd = 2.83, sd_digits = 2, Z = "integer")

# an SD too small to be possible: caught by the bounds test
sd_bounds_check(l = 1, u = 7, n = 30, mean = 2.97, mean_digits = 2,
                sd = 0.10, sd_digits = 2, Z = "quasiinteger")
```

The check also returns percent-of-maximum-possible (POMP) transforms:
`pomp_mean`, a linear `pomp_sd_parity` (SD as a fraction of the mean-agnostic
maximum), and `pomp_sd_sharp` (the SD's position within the sharp
mean-conditional band: 0 at the floor, 1 at the ceiling, outside `[0, 1]` when
impossible).

## Across a data frame

`sd_bounds_check_multiple()` applies the check to each row of a data frame.
Columns named like the arguments are read per row; anything passed as a constant
is broadcast. Repeated constraint tuples are computed once and reused.

```{r}
reports <- data.frame(
  mean = c(2.97, 4.50, 3.10),
  sd   = c(2.83, 0.10, 1.90)
)
sd_bounds_check_multiple(reports, l = 1, u = 7, n = 30,
                         mean_digits = 2, sd_digits = 2, Z = "integer")[
  , c("mean", "sd", "consistent", "failed_tests")]
```

## Visualising the feasible envelope

`sd_bounds_curve()` traces the floor and ceiling across the mean;
`plot_sd_bounds()` draws the band and, optionally, reported points coloured by
consistency.

```{r}
curve  <- sd_bounds_curve(l = 1, u = 7, n = 30, Z = "quasiinteger")
points <- sd_bounds_check_multiple(
  data.frame(mean = c(3.0, 4.0, 2.5), sd = c(2.5, 0.2, 3.6)),
  l = 1, u = 7, n = 30, mean_digits = 1, sd_digits = 1, Z = "quasiinteger")

plot_sd_bounds(curve, points = points)
```

On the POMP scale, `reference = "parity"` keeps the geometry undistorted (a
linear rescaling), while `reference = "sharp"` maps the feasible region to the
unit square so a point's height is its position within the band.

```{r}
plot_sd_bounds_pomp(curve, points = points, reference = "sharp")
```

## The umbrella

`umbrella_data()` builds the full grid of reported (mean, SD) pairs at a given
reporting precision, tagging each as consistent, GRIMMER-inconsistent, or out of
bounds; `plot_umbrella()` renders it.

```{r}
um <- umbrella_data(n = 14, l = 1, u = 7, digits = 2)
plot_umbrella(um, curve = sd_bounds_curve(l = 1, u = 7, n = 14, Z = "quasiinteger"))
```

## Multi-item scales and composites

`scoring` and `n_items` describe how response items form the score. Sum scores
are integer-gridded (`scoring = "sumscored"`); item means live on a finer
`1/n_items` grid (`scoring = "meanscored"`). A reported Cronbach's alpha
tightens the bounds of a composite further.

```{r}
# a 3-item composite reported as an item mean, with alpha
sd_bounds(l = 1, u = 5, n = 20, mean = 3, Z = "integer",
          scoring = "meanscored", n_items = 3, alpha = 0.8)
```
