---
title: "tides"
subtitle: "Checking reported summary statistics from truncated data"
author: "Ian Hussey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
options(scipen = 999) # disable scientific notation
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(tides)

# used in the vignette but not required by tides itself
library(scrutiny)
```

## What tides does

When a measure is bounded (truncated) — a Likert item, a symptom inventory, a
percentage — the reported mean constrains the standard deviation that is
arithmetically possible. `tides()` computes the smallest and largest SD that a
reported mean, sample size and scale range could have produced, and flags a
report as inconsistent when the reported SD falls outside that range.

We illustrate with the summary statistics reported by Zemestani & Mozaffari
(2020), where the outcome is the BDI-II depression inventory (possible scores
0–63).

## Summary statistics from the article

```{r}
# data manually extracted from the article
reported_results <- data.frame(
  timepoint = c(          "t1",           "t2",           "t3",      "t1",      "t2",      "t3"),
  condition = c("intervention", "intervention", "intervention", "control", "control", "control"),
  n         = c(         23.00,          23.00,          23.00,     29.00,     29.00,     29.00),
  mean      = c(         31.09,          20.70,          24.09,     31.92,     32.57,     33.40),
  sd        = c(          6.03,           3.40,           3.77,      4.83,      5.15,      4.45),
  min       = 0,   # BDI-II minimum score
  max       = 63,  # BDI-II maximum score
  n_items   = 1,
  digits    = 2
)

reported_results
```

## Running TIDES over a table

`tides_df()` applies `tides()` to each row, taking its arguments from the
columns of the same name.

```{r}
tides_results <- reported_results |>
  tides_df()

tides_results |>
  select(timepoint, condition, n, mean, sd,
         min_sd, max_sd, tides_consistent)
```

Every reported SD here sits inside its feasible `[min_sd, max_sd]` range, so
none of these values are TIDES-inconsistent.

## Comparing TIDES with GRIM and GRIMMER

TIDES is complementary to the granularity-based checks GRIM and GRIMMER (from
the `scrutiny` package). GRIM checks whether a reported mean is possible given
the sample size and number of items; GRIMMER extends this to the SD. TIDES adds
the constraint imposed by the scale bounds.

The `scrutiny` GRIM/GRIMMER interface has changed across releases (older
versions take zero-padded strings; newer ones take numbers plus their decimal
places), so we wrap each in a small helper that works with either.

```{r}
grim_ok <- function(x, n, digits, items = 1) {
  if ("digits_x" %in% names(formals(scrutiny::grim))) {
    scrutiny::grim(x = x, n = n, digits_x = digits, items = items)
  } else {
    scrutiny::grim(x = scrutiny::restore_zeros(x, width = digits),
                   n = n, items = items)
  }
}

grimmer_ok <- function(x, sd, n, digits, items = 1) {
  if ("digits_x" %in% names(formals(scrutiny::grimmer))) {
    scrutiny::grimmer(x = x, sd = sd, n = n,
                      digits_x = digits, digits_sd = digits, items = items)
  } else {
    scrutiny::grimmer(x = scrutiny::restore_zeros(x, width = digits),
                      sd = scrutiny::restore_zeros(sd, width = digits),
                      n = n, items = items)
  }
}

reported_results |>
  mutate(
    grim = pmap_lgl(list(x = mean, n = n, digits = digits, items = n_items),
                    grim_ok),
    grimmer = pmap_lgl(list(x = mean, sd = sd, n = n,
                            digits = digits, items = n_items),
                       grimmer_ok)
  ) |>
  tides_df() |>
  select(timepoint, condition, n, mean, sd,
         grim_consistent = grim,
         grimmer_consistent = grimmer,
         tides_consistent)
```

## Visualising the feasible envelope

For values that share the same sample size, scale range and precision,
`plot_tides()` draws the TIDES envelope with the reported points overlaid.
Here we plot the three intervention-group timepoints.

```{r}
reported_results |>
  filter(condition == "intervention") |>
  mutate(calculate_min_sd = TRUE, method = "approximate") |>
  tides_df() |>
  plot_tides()
```

## References

Brown, N. J. L., & Heathers, J. A. J. (2017). The GRIM test: A simple technique
detects numerous anomalies in the reporting of results in psychology. *Social
Psychological and Personality Science, 8*(4), 363–369.

Zemestani, M., & Mozaffari, S. (2020). Acceptance and commitment therapy for
the treatment of depression in persons with physical disability. *(as cited for
illustration only).*
