---
title: "Briding umbrella plots and tides"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---
  
```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

library(tidyverse)
library(janitor)
library(rsprite2)
library(scrutiny) 
library(purrr)
library(patchwork)
library(knitr)
library(kableExtra)

min_decimals <- function(x, digits = 2) {
  sprintf(paste0("%.", digits, "f"), x)
}

```

# Illustrate error in .sd_limits

This works correctly: all participants respond 1, therefore M = 1, SD = 0

```{r}

n_digits <- 2

dat <- 
  tibble(score = rep(1, times = 100)) |>
  summarize(mean = round_half_up(mean(score), n_digits),
            sd = round_half_up(sd(score), n_digits),
            n = n())

GRIMMER_test(mean = dat$mean, 
             sd = dat$sd, 
             n_obs = dat$n, 
             m_prec = n_digits, 
             sd_prec = n_digits, 
             n_items = 1, 
             min_val = 1, 
             max_val = 7)

```

bug: this should return TRUE but it returns an error. Akin to the above example, all participants respond 7, therefore M = 7, SD = 0.

Chunk set to do not run so to allow knit.

```{r eval=FALSE, include=TRUE}

dat <- 
  tibble(score = rep(7, times = 100)) |>
  summarize(mean = round_half_up(mean(score), n_digits),
            sd = round_half_up(sd(score), n_digits),
            n = n())

GRIMMER_test(mean = dat$mean, 
             sd = dat$sd, 
             n_obs = dat$n, 
             m_prec = n_digits, 
             sd_prec = n_digits, 
             n_items = 1, 
             min_val = 1, 
             max_val = 7) 

```

## original code for .sd_limits

```{r}

.sd_limits <- function(n_obs, mean, min_val, max_val, sd_prec = NULL, n_items = 1) {
  
  if (is.null(sd_prec)) {
    sd_prec <- max(nchar(sub("^[0-9]*", "", mean)) - 1, 0)
  }
  
  result <- c(-Inf, Inf)
  
  aMax <- min_val                                # "aMax" means "value of a to produce the max SD"
  aMin <- floor(mean*n_items)/n_items
  bMax <- max(max_val, min_val + 1, aMin + 1)   # sanity check (just max_val would normally be ok)
  bMin <- aMin + 1/n_items
  total <- round(mean * n_obs * n_items)/n_items
  
  poss_values <- max_val
  for (i in seq_len(n_items)) {
    poss_values <- c(poss_values, min_val:(max_val-1) + (1 / n_items) * (i - 1))
  }
  poss_values <- sort(poss_values)
  
  for (abm in list(c(aMin, bMin, 1), c(aMax, bMax, 2))) {
    
    a <- abm[1]
    b <- abm[2]
    m <- abm[3]
    
    
    k <- round((total - (n_obs * b)) / (a - b))
    k <- min(max(k, 1), n_obs - 1)               # ensure there is at least one of each of two numbers
    vec <- c(rep(a, k), rep(b, n_obs - k))
    diff <- sum(vec) - total
    
    if ((diff < 0)) {
      vec <- c(rep(a, k - 1), a + abs(diff), rep(b, n_obs - k))
    }
    else if ((diff > 0)) {
      vec <- c(rep(a, k), b - diff, rep(b, n_obs - k - 1))
    }
    
    # # Debugging: Print relevant information to diagnose the issue
    # cat("Iteration for a =", a, "b =", b, "\n")
    # cat("Generated vec:", vec, "\n")
    # cat("Mean of vec:", mean(vec), "Expected mean:", mean, "\n")
    # cat("SD of vec:", sd(vec), "\n\n")
    
    if (round(mean(vec), sd_prec) != round(mean, sd_prec) | !all(floor(vec*10e9) %in% floor(poss_values*10e9))) {
      stop("Error in calculating range of possible standard deviations")
    }
    
    result[m] <- round(sd(vec), sd_prec)
  }
  
  return(result)
}

```

## potential fix

```{r}

.sd_limits_fixed <- function(n_obs, mean, min_val, max_val, sd_prec = NULL, n_items = 1) {
  
  if (is.null(sd_prec)) {
    sd_prec <- max(nchar(sub("^[0-9]*", "", mean)) - 1, 0)
  }
  
  result <- c(-Inf, Inf)
  
  aMax <- min_val
  aMin <- floor(mean*n_items)/n_items
  # bMax <- max(max_val, min_val + 1, aMin + 1) # original
  bMax <- min(max(max_val, min_val + 1, aMin + 1), max_val)   # Adjusted here
  # bMin <- aMin + 1/n_items # original
  bMin <- min(aMin + 1/n_items, max_val)                      # Adjusted here
  total <- round(mean * n_obs * n_items)/n_items
  
  poss_values <- max_val
  for (i in seq_len(n_items)) {
    poss_values <- c(poss_values, min_val:(max_val-1) + (1 / n_items) * (i - 1))
  }
  poss_values <- sort(poss_values)
  
  for (abm in list(c(aMin, bMin, 1), c(aMax, bMax, 2))) {
    
    a <- abm[1]
    b <- abm[2]
    m <- abm[3]
    
    # Adjust a and b to be within min_val and max_val
    a <- min(max(a, min_val), max_val)
    b <- min(max(b, min_val), max_val)
   
    if (a == b) {
      vec <- rep(a, n_obs)
    } else {
      k <- round((total - (n_obs * b)) / (a - b))
      k <- min(max(k, 1), n_obs - 1)
      vec <- c(rep(a, k), rep(b, n_obs - k))
      diff <- sum(vec) - total
      
      if ((diff < 0)) {
        vec <- c(rep(a, k - 1), a + abs(diff), rep(b, n_obs - k))
      } else if ((diff > 0)) {
        vec <- c(rep(a, k), b - diff, rep(b, n_obs - k - 1))
      }
    }
    
    if (round(mean(vec), sd_prec) != round(mean, sd_prec) | !all(floor(vec*10e9) %in% floor(poss_values*10e9))) {
      stop("Error in calculating range of possible standard deviations")
    }
    
    result[m] <- round(sd(vec), sd_prec)
  }
  
  return(result)
}

# illustrate what i think is a bug in .sd_limits: 
n_digits <- 2

# works correctly
# all participants respond 1, therefore M = 1, SD = 0
dat <- 
  tibble(score = rep(1, times = 100)) |>
  summarize(mean = round_half_up(mean(score), n_digits),
            sd = round_half_up(sd(score), n_digits),
            n = n())

.sd_limits_fixed(n_obs = dat$n, 
                 mean = dat$mean, 
                 min_val = 1, 
                 max_val = 7,
                 sd_prec = n_digits, 
                 n_items = 1)


# works correctly
# half repond 1, half repond 7, therefore M = 4, SD approaches 3 as N approaches Inf, is slightly higher than 3 in small samples
dat <- 
  tibble(score = c(rep(1, times = 50), rep(7, times = 50))) |>
  summarize(mean = round_half_up(mean(score), n_digits),
            sd = round_half_up(sd(score), n_digits),
            n = n())

.sd_limits_fixed(n_obs = dat$n, 
                 mean = dat$mean, 
                 min_val = 1, 
                 max_val = 7,
                 sd_prec = n_digits, 
                 n_items = 1)


# previously threw an error but now doesn't
# all participants respond 7, therefore M = 7, SD = 0
dat <- 
  tibble(score = rep(7, times = 100)) |>
  summarize(mean = round_half_up(mean(score), n_digits),
            sd = round_half_up(sd(score), n_digits),
            n = n())

.sd_limits_fixed(n_obs = dat$n, 
                 mean = dat$mean, 
                 min_val = 1, 
                 max_val = 7,
                 sd_prec = n_digits, 
                 n_items = 1)

```

# Umbrella plot using original .sd_limits() illustrating problem

## Generate data

```{r}

# create possibly version of GRIMMER_test that fails quietly, given that it has a bug
possibly_GRIMMER_test = possibly(GRIMMER_test, otherwise = FALSE)

dat <- 
  expand_grid(mean = seq(from = 1, to = 7, by = 0.01),
              sd = seq(from = 0, to = 3.5, by = 0.01)) |>
  mutate(n_obs   = 14,
         m_prec  = 2,
         sd_prec = 2,
         n_items = 1,
         min_val = 1, 
         max_val = 7) |>
  mutate(grim    = pmap(list(mean, n_obs, m_prec, n_items), GRIM_test)) |>
  mutate(grimmer = pmap(list(mean, sd, n_obs, m_prec, sd_prec, n_items, min_val, max_val), possibly_GRIMMER_test)) |>
  unnest(grim) |>
  unnest(grimmer)
#filter(grim & grimmer)

```

## Plot

```{r fig.height=7, fig.width=10}

# only GRIM-consistent values
dat |>
  filter(grim) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

# only GRIMMER-consistent values
dat |>
  filter(grimmer) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

# only GRIM and GRIMMER-consistent values
dat |>
  filter(grim & grimmer) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 1, color = "grey60") +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

```

- observe the vertical gaps in the plot, and the original bug of the point at (7,0) being missing.
- some of these are due to issues with .sd_limits() eg the point at (7,0), and others seemt to be issues with rsprite2's::GRIMMER_test implementation eg the missing column of values at 2.86. 

# Attempt 1 to fix Umbrella plot

## Generate data

```{r}

GRIMMER_test_fixed <- function(mean, sd, n_obs, m_prec = NULL, sd_prec = NULL, n_items = 1, min_val = NULL, max_val = NULL) {
  if (is.null(m_prec)) {
    m_prec <- max(nchar(sub("^[0-9]*", "", mean)) - 1, 0)
  }
  
  if (is.null(sd_prec)) {
    sd_prec <- max(nchar(sub("^[0-9]*", "", sd)) - 1, 0)
  }
  
  assert_count(m_prec)
  assert_count(sd_prec)
  assert_count(n_obs)
  assert_count(n_items)
  assert_number(mean)
  assert_number(sd)
  
  effective_n = n_obs * n_items
  
  # Applies the GRIM test, and computes the possible mean.
  sum <- mean * effective_n
  realsum <- round(sum)
  realmean <- realsum / effective_n
  
  #Checks whether mean and SD are within possible range
  if (!is.null(min_val) & !is.null(max_val)) {
    if (mean < min_val | mean > max_val) {
      warning("The mean must be between the scale minimum and maximum")
      return(FALSE)
    }
    sd_limits <- .sd_limits_fixed(n_obs, mean, min_val, max_val, sd_prec, n_items)
    if (sd < sd_limits[1] | sd > sd_limits[2]) {
      warning("Given the scale minimum and maximum, the standard deviation has to be between ", sd_limits[1], " and ", sd_limits[2], ".")
      return(FALSE)
    }
  }
  # Creates functions to round a number consistently up or down, when the last digit is 5
  round_down <- function(number, decimals = 2) {
    to_round <- number * 10^(decimals + 1) - floor(number * 10^(decimals)) * 10
    number_rounded <- ifelse(to_round == 5,
                             floor(number * 10^decimals) / 10^decimals,
                             round(number, digits = decimals))
    return(number_rounded)
  }
  
  round_up <- function(number, decimals = 2) {
    to_round <- number * 10^(decimals + 1) - floor(number * 10^(decimals)) * 10
    number_rounded <- ifelse(to_round == 5,
                             ceiling(number * 10^decimals) / 10^decimals,
                             round(number, digits = decimals))
    return(number_rounded)
  }
  
  # Applies the GRIM test, to see whether the reconstituted mean is the same as the reported mean (with both down and up rounding)
  
  consistent_down <- round_down(number = realmean, decimals = m_prec) == mean
  consistent_up <- round_up(number = realmean, decimals = m_prec) == mean
  
  if (!consistent_down & !consistent_up) {
    warning("GRIM inconsistent - so GRIMMER test cannot be run. See ?GRIM_test")
    return(FALSE)
  }
  
  # Computes the lower and upper bounds for the sd.
  
  Lsigma <- ifelse(sd < 5 / (10^(sd_prec+1)), 0, sd - 5 / (10^(sd_prec+1)))
  Usigma <- sd + 5 / (10^(sd_prec+1))
  
  # Computes the lower and upper bounds for the sum of squares of items.
  
  lower_bound <- ((n_obs - 1) * Lsigma^2 + n_obs * realmean^2)*n_items^2
  upper_bound <- ((n_obs - 1) * Usigma^2 + n_obs * realmean^2)*n_items^2
  
  # Checks that there is at least an integer between the lower and upperbound
  
  if (ceiling(lower_bound) > floor(upper_bound)) {
    return(FALSE)
  }
  
  # Takes a vector of all the integers between the lowerbound and upperbound
  
  possible_integers <- ceiling(lower_bound):floor(upper_bound)
  
  # Creates the predicted variance and sd
  
  Predicted_Variance <- (possible_integers/n_items^2 - n_obs * realmean^2) / (n_obs - 1)
  Predicted_SD <- sqrt(Predicted_Variance)
  
  # Computes whether one Predicted_SD matches the SD (trying to round both down and up)
  
  Rounded_SD_down <- round_down(Predicted_SD, sd_prec)
  Rounded_SD_up <- round_up(Predicted_SD, sd_prec)
  
  Matches_SD <- Rounded_SD_down == sd | Rounded_SD_up == sd
  
  if (!any(Matches_SD)) {
    return(FALSE)
  }
  
  # Computes whether there is an integer of the correct oddness between the lower and upper bounds.
  oddness <- realsum %% 2
  Matches_Oddness <- possible_integers %% 2 == oddness
  return(any(Matches_SD & Matches_Oddness))
  
  return(TRUE)
}

library(checkmate) # dependency not specified by rsprite2??

GRIMMER_test_fixed(mean = 1.00,
                   sd = 0.00,
                   n_obs = 14,
                   m_prec = 2,
                   sd_prec = 2,
                   n_items = 1,
                   min_val = 1,
                   max_val = 7)


GRIMMER_test_fixed(mean = 2.10,
                   sd = 0.90,
                   n_obs = 31,
                   m_prec = 2,
                   sd_prec = 2,
                   n_items = 1,
                   min_val = 1,
                   max_val = 7)

# create possibly version of GRIMMER_test that fails quietly, given that it has a bug
possibly_GRIMMER_test_fixed = possibly(GRIMMER_test_fixed, otherwise = FALSE)

dat <- 
  expand_grid(mean = seq(from = 1, to = 7, by = 0.01),
              sd = seq(from = 0, to = 3.5, by = 0.01)) |>
  mutate(n_obs   = 14,
         m_prec  = 2,
         sd_prec = 2,
         n_items = 1,
         min_val = 1, 
         max_val = 7) |>
  mutate(grim    = pmap(list(mean, n_obs, m_prec, n_items), GRIM_test)) |>
  mutate(grimmer = pmap(list(mean, sd, n_obs, m_prec, sd_prec, n_items, min_val, max_val), possibly_GRIMMER_test_fixed)) |>
  unnest(grim) |>
  unnest(grimmer)
#filter(grim & grimmer)

```

## Plot

```{r fig.height=7, fig.width=10}

# # only GRIM-consistent values
# dat |>
#   filter(grim) |>
#   ggplot(aes(mean, sd)) +
#   geom_point(shape = 15, size = 1, color = "grey60") +
#   theme_linedraw() +
#   scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
#   scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))
# 
# # only GRIMMER-consistent values
# dat |>
#   filter(grimmer) |>
#   ggplot(aes(mean, sd)) +
#   geom_point(shape = 15, size = 1, color = "grey60") +
#   theme_linedraw() +
#   scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
#   scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

# only GRIM and GRIMMER-consistent values
dat |>
  filter(grim & grimmer) |>
  ggplot(aes(mean, sd)) +
  geom_point(shape = 15, size = 0.7, color = "grey0", alpha = 0.4) +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 3.5)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(1, 7))

# temp <- dat |>
#   filter(grim & grimmer) |>
#   filter(mean > 2.7 & mean < 3.1)
# 
# writexl::write_xlsx(temp, "temp.xlsx")

```

\TODO REDO THIS WITH scrutiny::grimmer()

- point at (7,0) now appears, but vertical columns of values that should bre present are still missing, eg at 2.86. the issue is with GRIMMER_test().

```{r}

# these values pass grimmer on the web app http://www.prepubmed.org/grimmer_sd/?type=unknown&direction=Up&sd=2.17&mean=2.86&size=14
# but fail GRIMMER_test
GRIMMER_test_fixed(mean = 2.86,
                   sd = 2.17,
                   n_obs = 14,
                   m_prec = 2,
                   sd_prec = 2,
                   n_items = 1,
                   min_val = 1,
                   max_val = 7)

library(scrutiny) 

grimmer(x = "2.86",
        sd = "2.17",
        n = 14,
        items = 1)

```

# GRIM vs GRIM+GRIMMER vs GRIM+GRIMMER+Umbrella plot using scrutiny::grimmer + tides_modified_from_.sd_limits()

```{r}

tides_modified_from_.sd_limits <- function(n_obs, mean, sd, min_val, max_val, sd_prec = NULL, n_items = 1) {
  
  if (is.null(sd_prec)) {
    sd_prec <- max(nchar(sub("^[0-9]*", "", mean)) - 1, 0)
  }
  
  result <- c(-Inf, Inf)
  
  aMax <- min_val
  aMin <- floor(mean*n_items)/n_items
  bMax <- min(max(max_val, min_val + 1, aMin + 1), max_val)   # Adjusted here
  bMin <- min(aMin + 1/n_items, max_val)                      # Adjusted here
  total <- round(mean * n_obs * n_items)/n_items
  
  poss_values <- max_val
  for (i in seq_len(n_items)) {
    poss_values <- c(poss_values, min_val:(max_val-1) + (1 / n_items) * (i - 1))
  }
  poss_values <- sort(poss_values)
  
  for (abm in list(c(aMin, bMin, 1), c(aMax, bMax, 2))) {
    
    a <- abm[1]
    b <- abm[2]
    m <- abm[3]
    
    # Adjust a and b to be within min_val and max_val
    a <- min(max(a, min_val), max_val)
    b <- min(max(b, min_val), max_val)
   
    if (a == b) {
      vec <- rep(a, n_obs)
    } else {
      k <- round((total - (n_obs * b)) / (a - b))
      k <- min(max(k, 1), n_obs - 1)
      vec <- c(rep(a, k), rep(b, n_obs - k))
      diff <- sum(vec) - total
      
      if ((diff < 0)) {
        vec <- c(rep(a, k - 1), a + abs(diff), rep(b, n_obs - k))
      } else if ((diff > 0)) {
        vec <- c(rep(a, k), b - diff, rep(b, n_obs - k - 1))
      }
    }
    
    # instead of throwing errors here, return NA
    # if (round(mean(vec), sd_prec) != round(mean, sd_prec) | !all(floor(vec*10e9) %in% floor(poss_values*10e9))) {
    #   stop("Error in calculating range of possible standard deviations")
    # }
    # result[m] <- round(sd(vec), sd_prec)
    
    # Check if the calculated mean and values match expected conditions
    if (round(mean(vec), sd_prec) == round(mean, sd_prec) & all(floor(vec*10e9) %in% floor(poss_values*10e9))) {
      result[m] <- round(sd(vec), sd_prec)
    }
    
  }
  
  # Replace Inf or -Inf with NA
  result[is.infinite(result)] <- NA
  
  # returns df instead of vector
  res <- 
    data.frame(sd_min = result[1],
               sd_max = result[2]) |>
    mutate(tides = case_when(sd < sd_min ~ FALSE,
                             is.na(sd_min) ~ FALSE,
                             sd > sd_max ~ FALSE,
                             is.na(sd_max) ~ FALSE,
                             TRUE ~ TRUE))
  
  return(res)
}

# dat_n_11 <-
#   expand_grid(mean = seq(from = 0.5, to = 7.5, by = 0.01),
#               sd = seq(from = 0, to = 4, by = 0.01)) |>
#   mutate(n_obs   = 11,
#          m_prec  = 2,
#          sd_prec = 2,
#          n_items = 1,
#          min_val = 1,
#          max_val = 7) |>
#   mutate(grim = pmap(list(as.character(mean), n_obs),
#                      grim)) |>
#   mutate(grimmer = pmap(list(as.character(mean), as.character(sd), n_obs),
#                         grimmer)) |>
#   mutate(tides = pmap(list(n_obs, mean, sd, min_val, max_val, sd_prec, n_items),
#                       tides_modified_from_.sd_limits)) |>
#   unnest(grim) |>
#   unnest(grimmer) |>
#   unnest(tides)
# 
# dat_n_100 <-
#   expand_grid(mean = seq(from = 0.5, to = 7.5, by = 0.01),
#               sd = seq(from = 0, to = 4, by = 0.01)) |>
#   mutate(n_obs   = 100,
#          m_prec  = 2,
#          sd_prec = 2,
#          n_items = 1,
#          min_val = 1,
#          max_val = 7) |>
#   mutate(grim = pmap(list(as.character(mean), n_obs),
#                      grim)) |>
#   mutate(grimmer = pmap(list(as.character(mean), as.character(sd), n_obs),
#                         grimmer)) |>
#   mutate(tides = pmap(list(n_obs, mean, sd, min_val, max_val, sd_prec, n_items),
#                       tides_modified_from_.sd_limits)) |>
#   unnest(grim) |>
#   unnest(grimmer) |>
#   unnest(tides)
# 
# write_rds(dat_n_11, "results_grim_grimmer_tides_n_11.rds")
# write_rds(dat_n_100, "results_grim_grimmer_tides_n_100.rds")

dat_n_11 <- read_rds("results_grim_grimmer_tides_n_11.rds")
dat_n_100 <- read_rds("results_grim_grimmer_tides_n_100.rds")

```

## Plot N=11

```{r fig.height=16, fig.width=7}

dat_n_11 |>
  count()

dat_n_11 |>
  count(grim, grimmer, tides)

dat_n_11 <- dat_n_11 |>
  mutate(label = case_when(mean < 1 | mean > 7 ~ "Impossible score flagged as possible",
                           sd < 0 | sd > 3.6 ~ "Impossible score flagged as possible", # Croucher (2004) loose heuristic of max SD as 60% of range (ie 6 * .6 = .36)
                           TRUE ~ "Possible score flagged as possible"))

# only GRIM-consistent values
p_grim_n_11 <- dat_n_11 |>
  filter(grim) |>
  # mutate(label = case_when(grim == FALSE ~ "Impossible score flagged as impossible",
  #                          mean < 1 | mean > 7 ~ "Impossible score flagged as possible",
  #                          sd < 0 | sd > 3.6 ~ "Impossible score flagged as possible", # Croucher (2004) loose heuristic of max SD as 60% of range (ie 6 * .6 = .36)
  #                          TRUE ~ "Possible score flagged as possible")) |>
  ggplot(aes(mean, sd, color = label)) +
  geom_point(shape = 15, size = 0.5) + # "grey20"
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), 
                     limit = c(0, 4), 
                     expand = c(0.01, 0.01)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), 
                     limit = c(0.5, 7.5), 
                     expand = c(0.01, 0.01)) +
  scale_color_viridis_d(begin = 0.3, end = 0.7) +
  ylab("Standard Deviation") +
  xlab("Mean") +
  ggtitle("GRIM consistent values") +
  theme(legend.position = "top",
        legend.direction = "vertical") +
  guides(color = guide_legend(override.aes = list(size = 4, ncol = 1), title = NULL))

# only GRIM and GRIMMER-consistent values
p_grimmer_n_11 <- dat_n_11 |>
  filter(grim & grimmer) |>
  # mutate(label = case_when(grim == FALSE | grimmer == FALSE ~ "Impossible score flagged as impossible",
  #                          mean < 1 | mean > 7 ~ "Impossible score flagged as possible",
  #                          sd < 0 | sd > 3.6 ~ "Impossible score flagged as possible", # Croucher (2004) loose heuristic of max SD as 60% of range (ie 6 * .6 = .36)
  #                          TRUE ~ "Possible score flagged as possible")) |>
  ggplot(aes(mean, sd, color = label)) +
  geom_point(shape = 15, size = 0.5) +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), 
                     limit = c(0, 4), 
                     expand = c(0.01, 0.01)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), 
                     limit = c(0.5, 7.5), 
                     expand = c(0.01, 0.01)) +
  scale_color_viridis_d(begin = 0.3, end = 0.7) +
  ylab("Standard Deviation") +
  xlab("Mean") +
  ggtitle("GRIM + GRIMMER consistent values") +
  theme(legend.position = "none")

# only GRIM and GRIMMER and TIDES-consistent values
p_tides_n_11 <- dat_n_11 |>
  filter(grim & grimmer & tides) |>
  # mutate(label = case_when(grim == FALSE | grimmer == FALSE | tides == FALSE ~ "Impossible score flagged as impossible",
  #                          mean < 1 | mean > 7 ~ "Impossible score flagged as possible",
  #                          sd < 0 | sd > 3.6 ~ "Impossible score flagged as possible", # Croucher (2004) loose heuristic of max SD as 60% of range (ie 6 * .6 = .36)
  #                          TRUE ~ "Possible score flagged as possible")) |>
  ggplot(aes(mean, sd, color = label)) +
  geom_point(shape = 15, size = 0.5) +
  theme_linedraw() +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8), 
                     limit = c(0, 4), 
                     expand = c(0.01, 0.01)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 7), 
                     limit = c(0.5, 7.5), 
                     expand = c(0.01, 0.01)) +
  scale_color_viridis_d(begin = 0.3, end = 0.7, direction = -1) +
  ylab("Standard Deviation") +
  xlab("Mean") +
  ggtitle("GRIM + GRIMMER + TIDES consistent values") +
  theme(legend.position = "none")
  

# dat_n_11 |>
#   filter(grim & grimmer & tides) |>
#   group_by(mean) |>
#   filter(sd == max(sd) | sd == min(sd)) |>
#   mutate(group = ifelse(sd == max(sd), "max", "min")) |>
#   ungroup() |>
#   ggplot(aes(mean, sd, group = group)) +
#   #geom_point(shape = 15, size = 1, color = "grey20") +
#   geom_line() +
#   theme_linedraw() +
#   scale_y_continuous(breaks = scales::breaks_pretty(n = 8), limit = c(0, 4)) +
#   scale_x_continuous(breaks = scales::breaks_pretty(n = 7), limit = c(0.5, 7.5))

p_grim_n_11 + p_grimmer_n_11 + p_tides_n_11 + plot_layout(ncol = 1)

```






