---
title: "Validating `brimmer()` against `brimmest()`"
subtitle: "Every verdict on whole reporting grids, judged against an exact oracle. `brimmer()` is applied grid-wise via `umbrella_data()`, whose equivalence to it is checked first."
author: "Ian Hussey"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

<!--
Companion to validate_sd_bounds_functions.qmd. That document validates claims
about individual functions; this one validates the VERDICTS brimmer() returns,
over whole reporting grids, against brimmest() as an exact oracle.

brimmer() is exercised grid-wise through umbrella_data(), which applies the same
three tests (SD bounds, GRIM, GRIMMER) to every cell of a reporting grid in one
pass. Calling brimmer() per tuple over ~5.6e5 cells would take hours, so the
substitution is made deliberately and its exactness is asserted before the sweep
(see "brimmer() and the grid builder agree"). Witness datasets later in the
document call brimmer() directly.

The claim being verified is one-sided, and is the only one that matters for a
forensic tool: brimmer() must never call a report impossible when real integer
data can produce it. Everything else here is a measurement, not a guarantee.
-->

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)
library(knitr)
library(scrutiny)

has_unsum <- requireNamespace("unsum", quietly = TRUE)

# Dev-time document: load the package from the repo root. verification/ is
# excluded from the build via .Rbuildignore, so this never runs under R CMD check.
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..", quiet = TRUE)
} else {
  library(strait)
}

dir.create("models", showWarnings = FALSE)

# every hard claim in this document is registered here and asserted at the end
checks <- list()
check <- function(ok, label) { checks[[label]] <<- isTRUE(ok); invisible(ok) }

cached <- function(file, expr) {
  path <- file.path("models", file)
  if (file.exists(path)) readRDS(path) else { x <- force(expr); saveRDS(x, path); x }
}
```

# What is verified, and against what

`brimmer()` screens a reported (mean, SD, *n*) triple with three tests: the
closed-form SD bounds, plus GRIM and GRIMMER, which are deferred to `scrutiny`.
Each is **necessary** — failing one proves the report impossible — and none is
sufficient.

Throughout the sweeps below, `brimmer()` is applied through `umbrella_data()`,
which runs those same three tests across an entire reporting grid in one pass.
The two are shown to return identical verdicts before any sweep is run; the
witness datasets at the end call `brimmer()` directly.

`brimmest()` is exact in both directions. It enumerates the reachable
*(sum, sum of squares)* states for the design by dynamic programming, rounds the
resulting lattice to the reporting precision, and tests membership. So for any
reported tuple it answers *possible* or *impossible* outright, and can serve as
the oracle the screen is measured against.

That gives two quantities per tuple, and they are not symmetric:

- a **false flag** — screen says impossible, `brimmest()` says possible — is a
  fatal error. It accuses real data.
- a **blind-spot cell** — screen says possible, `brimmest()` says impossible — is
  the expected price of a necessary-only test. It is a missed detection, not a
  false accusation.

This document sweeps both across designs. The false-flag count is asserted to be
zero everywhere; the blind spot is measured.

## Verifying the oracle first

Anchoring a sweep on `brimmest()` is only meaningful if `brimmest()` is itself
right, so it is checked against two independent constructions before use.

```{r oracle-brute}
# (i) brute force: every integer sample, no theory at all
brute_pairs <- function(n, l, u, digits) {
  g <- as.matrix(expand.grid(rep(list(l:u), n)))
  m <- cbind(rowMeans(g), apply(g, 1, stats::sd))
  # a reported pair is achievable if some exact pair rounds to it, under either
  # convention, matching brimmest()'s default rule set
  up   <- cbind(round_up(m[, 1], digits),   round_up(m[, 2], digits))
  down <- cbind(round_down(m[, 1], digits), round_down(m[, 2], digits))
  unique(rbind(up, down))
}

brute_ok <- map_lgl(list(c(4, 1, 5, 1), c(5, 0, 3, 1), c(6, 1, 4, 1)), function(cfg) {
  n <- cfg[1]; l <- cfg[2]; u <- cfg[3]; d <- cfg[4]
  b <- brute_pairs(n, l, u, d)
  grid <- expand_grid(mean = round(seq(l, u, by = 10^-d), d),
                      sd   = round(seq(0, (u - l), by = 10^-d), d))
  cert <- brimmest(l = l, u = u, n = n, mean = grid$mean, sd = grid$sd, digits = d)
  bk <- paste(sprintf("%.6f", b[, 1]), sprintf("%.6f", b[, 2]))
  gk <- paste(sprintf("%.6f", grid$mean), sprintf("%.6f", grid$sd))
  identical(sort(gk[cert$possible]), sort(intersect(gk, bk)))
})

check(all(brute_ok), "brimmest() == brute-force enumeration (3 small designs)")
tibble(design = c("n=4 on 1-5", "n=5 on 0-3", "n=6 on 1-4"), agrees = brute_ok) |>
  kable(caption = "brimmest() against exhaustive enumeration of every integer sample.")
```

```{r oracle-closure, eval=has_unsum}
# (ii) CLOSURE: an independent reconstruction algorithm in another package
n_c <- 9; l_c <- 1; u_c <- 5; d_c <- 1
um_c <- umbrella_data(n = n_c, l = l_c, u = u_c, digits = d_c, granularity = "integer")
cert_c <- brimmest(l = l_c, u = u_c, n = n_c, mean = um_c$mean, sd = um_c$sd,
                  digits = d_c)

closure_says <- cached("closure_n9_1to5.rds", {
  vapply(seq_len(nrow(um_c)), function(i) {
    r <- try(suppressWarnings(unsum::closure_generate(
      mean = sprintf("%.1f", um_c$mean[i]), sd = sprintf("%.1f", um_c$sd[i]),
      n = n_c, scale_min = l_c, scale_max = u_c, rounding = "up_or_down")),
      silent = TRUE)
    if (inherits(r, "try-error")) FALSE else nrow(r$results) > 0
  }, logical(1))
})

check(all(closure_says == cert_c$possible), "brimmest() == unsum::CLOSURE, cell for cell")
tibble(cells = nrow(um_c),
       agree = sum(closure_says == cert_c$possible),
       disagree = sum(closure_says != cert_c$possible)) |>
  kable(caption = "brimmest() against CLOSURE reconstruction.")
```

## brimmer() and the grid builder agree

The sweep below uses `umbrella_data()`, which applies the same three tests over a
whole grid in one pass, because calling `brimmer()` per tuple would make the
sweep hours long. That substitution is only legitimate if the two return the same
verdict, so it is checked directly on a random sample of the grid, including
cells whose mean fails GRIM.

```{r brimmer-equivalence}
set.seed(1)
l_e <- 1; u_e <- 5; n_e <- 12; d_e <- 1

um_e <- umbrella_data(n = n_e, l = l_e, u = u_e, digits = d_e, granularity = "integer")
grid_e <- expand_grid(mean = round(seq(l_e, u_e, by = 10^-d_e), d_e),
                      sd   = round(seq(0, 2.1, by = 10^-d_e), d_e)) |>
  left_join(um_e[, c("mean", "sd", "consistent")], by = c("mean", "sd")) |>
  mutate(grid_verdict = coalesce(consistent, FALSE)) |>
  slice_sample(n = 300)

grid_e$brimmer_verdict <- vapply(seq_len(nrow(grid_e)), function(i) {
  isTRUE(brimmer(l = l_e, u = u_e, n = n_e,
                 mean = grid_e$mean[i], digits_mean = d_e,
                 sd = grid_e$sd[i], digits_sd = d_e, granularity = "integer")$consistent)
}, logical(1))

check(all(grid_e$grid_verdict == grid_e$brimmer_verdict),
      "umbrella_data() verdict == brimmer() verdict (300 sampled tuples)")

tibble(sampled = nrow(grid_e),
       agree = sum(grid_e$grid_verdict == grid_e$brimmer_verdict),
       disagree = sum(grid_e$grid_verdict != grid_e$brimmer_verdict)) |>
  kable(caption = "The grid builder stands in for brimmer() exactly.")
```

# The sweep

Thirteen designs spanning three scales, sample sizes from 8 to 30, and one and
two decimal places of reporting precision. For each, the full reported grid is
built, every cell is screened and certified, and the two error directions are
counted.

```{r sweep}
designs <- bind_rows(
  expand_grid(l = 1, u = 5, n = c(8, 12, 20, 30), digits = 1),
  expand_grid(l = 1, u = 7, n = c(10, 20, 30),    digits = 1),
  expand_grid(l = 0, u = 6, n = c(10, 20),        digits = 1),
  expand_grid(l = 1, u = 5, n = c(10, 15),        digits = 2),
  expand_grid(l = 1, u = 7, n = c(12, 20),        digits = 2)
)

sweep_one <- function(l, u, n, digits) {
  step <- 10^-digits
  # the full reported grid, including means GRIM rejects (umbrella_data drops
  # those, so they must be reinstated to attribute the catch correctly)
  ceiling_sd <- sd_bounds(l = l, u = u, n = n)$max_sd
  grid <- expand_grid(mean = round(seq(l, u, by = step), digits),
                      sd   = round(seq(0, ceiling(ceiling_sd / step) * step, by = step), digits))

  um <- umbrella_data(n = n, l = l, u = u, digits = digits, granularity = "integer") |>
    select(mean, sd, in_bounds, grimmer, consistent)

  g <- grid |>
    left_join(um, by = c("mean", "sd")) |>
    # a cell absent from the umbrella grid had a GRIM-infeasible mean
    mutate(grim      = !is.na(consistent),
           in_bounds = coalesce(in_bounds, FALSE),
           grimmer   = coalesce(grimmer, FALSE),
           screened  = coalesce(consistent, FALSE),
           certified = brimmest(l = l, u = u, n = n, mean = mean, sd = sd,
                               digits = digits)$possible)

  ff <- g$certified & !g$screened          # false flags, by the test responsible
  tibble(l = l, u = u, n = n, digits = digits,
         cells       = nrow(g),
         screened    = sum(g$screened),
         certified   = sum(g$certified),
         false_flags = sum(ff),
         # strait's OWN test is the bounds; GRIM and GRIMMER are scrutiny's
         ff_bounds   = sum(ff & g$grim & !g$in_bounds),
         ff_grim     = sum(ff & !g$grim),
         ff_grimmer  = sum(ff & g$grim & g$in_bounds & !g$grimmer),
         blind_spot  = sum(g$screened & !g$certified),
         # which test catches each impossible cell
         caught_grim    = sum(!g$certified & !g$grim),
         caught_bounds  = sum(!g$certified & g$grim & !g$in_bounds),
         caught_grimmer = sum(!g$certified & g$grim & g$in_bounds & !g$grimmer))
}

sweep <- cached("sweep.rds", pmap_dfr(designs, sweep_one))

check(sum(sweep$ff_bounds) == 0, "the SD bounds never false-flag (single item)")
check(sum(sweep$false_flags) == 0, "no false flags at all in the single-item sweep")

sweep |>
  transmute(design = sprintf("%d-%d, n=%d, %ddp", l, u, n, digits),
            cells, screened, certified,
            `false flags` = false_flags,
            `blind spot` = blind_spot,
            `blind spot %` = sprintf("%.1f", 100 * blind_spot / pmax(screened, 1))) |>
  kable(caption = "The sweep. `false flags` is the fatal direction and is zero throughout.")
```

## The guarantee, and its direction

```{r direction}
tibble(
  quantity = c("cells swept",
               "certified possible",
               "screen rejected a certified-possible report (FATAL)",
               "screen admitted a certified-impossible report (missed detection)"),
  count = c(sum(sweep$cells), sum(sweep$certified),
            sum(sweep$false_flags), sum(sweep$blind_spot))
) |>
  kable(caption = "Aggregate over all designs.")
```

Across `r format(sum(sweep$cells), big.mark = ",")` reported tuples the screen
never once rejected a report that real integer data can produce. That is the
property the tool is for: a `brimmer()` failure is a proof of impossibility, and
can be acted on. A `brimmer()` pass is not a proof of possibility, and
`r format(sum(sweep$blind_spot), big.mark = ",")` of the passes are in fact
impossible — which is what `brimmest()` exists to settle.

## Which test earns the catch

Decomposing the impossible cells by the first test that rejects them shows how
much each component contributes, and how much neither does.

```{r attribution}
sweep |>
  transmute(design = sprintf("%d-%d, n=%d, %ddp", l, u, n, digits),
            impossible = cells - certified,
            GRIM = caught_grim, bounds = caught_bounds, GRIMMER = caught_grimmer,
            `caught by none (blind spot)` = blind_spot) |>
  kable(caption = "Attribution of impossible cells to the test that catches them.")
```

The bounds and the granularity tests are working on different axes, as the
manuscript argues: neither dominates, and the residual that neither catches is
the blind spot.

# How the blind spot scales

The blind spot is not a fixed overhead. GRIM's resolving power decays once
*n* exceeds `10^d`, and GRIMMER's detectable SD band narrows as `10^d / (n - 1)`,
so the share of screened-possible reports that are actually impossible grows
with *n*.

```{r prevalence}
prev_designs <- expand_grid(l = 1, u = 5, n = c(10, 20, 40, 80), digits = 2)
prevalence <- cached("prevalence.rds", pmap_dfr(prev_designs, sweep_one))

prevalence |>
  transmute(n, screened, blind_spot,
            `blind spot % of passes` = sprintf("%.1f", 100 * blind_spot / screened)) |>
  kable(caption = "Blind-spot prevalence on a 1-5 scale at 2 dp.")

# recorded, not asserted: the share depends on the scrutiny build (see below)
```

```{r prevalence-plot, fig.width = 6, fig.height = 3.4}
prevalence |>
  mutate(pct = 100 * blind_spot / screened) |>
  ggplot(aes(n, pct)) +
  geom_line(colour = "#1d4ed8") +
  geom_point(size = 2) +
  scale_x_continuous(breaks = prev_designs$n) +
  labs(x = "n", y = "blind spot, % of screened-possible reports",
       title = "The screen's misses grow with sample size") +
  theme_minimal(base_size = 11)
```

The share does not rise monotonically, and the shape is worth reading carefully.
The spike at *n* = 40 is not a property of the bounds: 40 is one of the sample
sizes affected by `scrutiny` issue `#86`, where floating-point error in the
candidate-sum range admits phantom sums, so GRIMMER passes cells it should
reject and they land in the blind spot. The measurement is therefore partly a
measurement of the installed `scrutiny` build, which is why it is recorded
rather than asserted.

The broader point survives regardless: a screen made of necessary conditions
leaves a residue at every *n*, so the value of settling survivors exactly with
`brimmest()` does not go away as designs get larger.

# Rounding conventions

A certificate of *possibility* is absolute: some sample rounds to the report
under the named rule. A certificate of *impossibility* holds only relative to the
rounding rules admitted. Narrowing the rule set therefore risks false flags, and
the sweep must show that.

```{r rounding}
round_check <- function(rules) {
  n <- 12; l <- 0; u <- 6; d <- 1
  um <- umbrella_data(n = n, l = l, u = u, digits = d, granularity = "integer")
  ce <- brimmest(l = l, u = u, n = n, mean = um$mean, sd = um$sd, digits = d,
                rounding = rules)
  tibble(rules = paste(rules, collapse = " + "),
         certified = sum(ce$possible))
}

bind_rows(round_check("half_up"), round_check("half_down"),
          round_check(c("half_up", "half_down"))) |>
  kable(caption = "Admitting fewer rounding conventions certifies fewer reports as possible.")
```

```{r rounding-witness}
# a real sample whose report is reachable ONLY when halves round down
x_w <- c(rep(0, 9), 1, 1, 1)
w_up   <- brimmest(l = 0, u = 6, n = 12, mean = 0.2, sd = 0.5, digits_mean = 1, digits_sd = 1, rounding = "half_up")
w_both <- brimmest(l = 0, u = 6, n = 12, mean = 0.2, sd = 0.5, digits_mean = 1, digits_sd = 1)

check(!w_up$possible && w_both$possible,
      "narrowing the rule set can wrongly reject a real sample")

tibble(sample = "c(rep(0, 9), 1, 1, 1)",
       mean = mean(x_w), sd = sd(x_w),
       `reported 1 dp` = "0.2, 0.5",
       `brimmest, half_up only` = w_up$possible,
       `brimmest, default rules` = w_both$possible) |>
  kable(digits = 4,
        caption = "Restricting rounding to a convention the source did not use is itself a false-flag risk.")
```

# Multi-item scoring

Composite scores move the data onto a `1/k` grid. `brimmest()` and the screen must
agree about that grid, in both sum-scored and mean-scored form.

```{r scoring}
scoring_one <- function(scoring, n_items, l, u, n, digits) {
  um <- umbrella_data(n = n, l = l, u = u, digits = digits, granularity = "integer",
                      scoring = scoring, n_items = n_items)
  ce <- brimmest(l = l, u = u, n = n, mean = um$mean, sd = um$sd, digits = digits,
                scoring = scoring, n_items = n_items)
  ff <- ce$possible & !um$consistent
  tibble(scoring = scoring, n_items = n_items,
         scale = sprintf("%d-%d", l, u), n = n,
         screened = sum(um$consistent),
         certified = sum(ce$possible),
         false_flags = sum(ff),
         ff_bounds  = sum(ff & !um$in_bounds),
         ff_grimmer = sum(ff & um$in_bounds & !um$grimmer),
         blind_spot = sum(um$consistent & !ce$possible))
}

scoring_grid <- bind_rows(
  tibble(scoring = "singleitem", n_items = 1, l = 1, u = 5,  n = 12, digits = 1),
  tibble(scoring = "sumscored",  n_items = 2, l = 0, u = 6,  n = 12, digits = 1),
  tibble(scoring = "sumscored",  n_items = 3, l = 3, u = 15, n = 10, digits = 1),
  tibble(scoring = "meanscored", n_items = 2, l = 1, u = 5,  n = 10, digits = 1),
  tibble(scoring = "meanscored", n_items = 3, l = 1, u = 7,  n = 8,  digits = 1)
)

scoring_res <- cached("scoring2.rds", pmap_dfr(scoring_grid, scoring_one))

check(sum(scoring_res$ff_bounds) == 0, "the SD bounds never false-flag in any scoring mode")

scoring_res |> kable(caption = "Screen against brimmest() across scoring modes. `ff_grimmer` is attributable to scrutiny, not to this package.")
```

## A false flag in scrutiny's multi-item GRIMMER

The scoring sweep is the one place a false flag appears, and it is not the
bounds. Every case has `in_bounds = TRUE` and `grimmer = FALSE`, so the report is
rejected by the deferred granularity test. A witness settles it: the report is
produced by a real sample.

```{r multiitem-witness}
x_mi <- c(2, 2, 2, 2, 2, 2, 2, 2.5, 2.5, 2.5)   # 10 people, each a mean of two 1-5 items

mi <- tibble(
  quantity = c("n", "exact mean", "exact SD", "reported at 1 dp",
               "brimmest(meanscored, k = 2)",
               "scrutiny grimmer(items = 2)",
               "scrutiny grimmer on the sum scale (items = 1)"),
  value = c(length(x_mi), sprintf("%.4f", mean(x_mi)), sprintf("%.4f", sd(x_mi)),
            "2.2, 0.2",
            as.character(brimmest(l = 1, u = 5, n = 10, mean = 2.2, sd = 0.2,
                                 digits_mean = 1, digits_sd = 1, scoring = "meanscored", n_items = 2)$possible),
            as.character(as.logical(grimmer(x = 2.2, sd = 0.2, n = 10, digits_x = 1,
                                            digits_sd = 1, items = 2))),
            as.character(as.logical(grimmer(x = 4.3, sd = 0.5, n = 10, digits_x = 1,
                                            digits_sd = 1, items = 1))))
)
mi |> kable(caption = "A real mean-scored sample that scrutiny's multi-item GRIMMER rejects.")

check(all(x_mi * 2 == round(x_mi * 2)), "witness lies on the 1/2 mean-score grid")
```

Every value is a legal 2-item mean, the reported pair is what a journal would
print, and `brimmest()` confirms attainability. `scrutiny::grimmer()` returns
`FALSE` under every rounding rule when told `items = 2`, while the same data
expressed on the sum scale passes. This is a **false flag in the `items`
handling**, distinct from the two issues already filed against `scrutiny`
(`#85`, the test-2/3 conjunction bug, and `#86`, floating-point boundary errors),
both of which are single-item problems.

Because it originates in a dependency and is version-dependent, it is recorded
here rather than asserted. What *is* asserted is that this package's own test —
the SD bounds — never contributes a false flag in any scoring mode.

# Witness datasets

The strongest form of the false-flag test needs no theory: construct real
samples, report them as a journal would, and require the screen to admit them.

```{r witnesses}
witness_designs <- list(
  list(x = c(rep(1, 7), rep(5, 5)),            l = 1, u = 5, d = 2),
  list(x = c(rep(0, 9), 1, 1, 1),              l = 0, u = 6, d = 1),
  list(x = c(1, rep(4, 10), 7),                l = 1, u = 7, d = 2),
  list(x = c(rep(2, 6), rep(3, 6)),            l = 1, u = 5, d = 2),
  list(x = c(rep(1, 13), 2),                   l = 1, u = 7, d = 2),
  list(x = c(rep(3, 19), 7),                   l = 1, u = 7, d = 2)
)

witnesses <- map_dfr(witness_designs, function(w) {
  n <- length(w$x)
  m_r <- round_up(mean(w$x), w$d); s_r <- round_up(sd(w$x), w$d)
  b <- brimmer(l = w$l, u = w$u, n = n, mean = m_r, digits_mean = w$d,
               sd = s_r, digits_sd = w$d, granularity = "integer")
  tibble(n = n, scale = sprintf("%d-%d", w$l, w$u),
         reported = sprintf("%.*f, %.*f", w$d, m_r, w$d, s_r),
         brimmer_consistent = isTRUE(b$consistent),
         certified = brimmest(l = w$l, u = w$u, n = n, mean = m_r, sd = s_r,
                             digits = w$d)$possible)
})

check(all(witnesses$certified), "every witness sample is certified possible")
check(all(witnesses$brimmer_consistent), "every witness sample passes brimmer()")

witnesses |> kable(caption = "Real samples, reported and re-screened. Any FALSE here is a false accusation.")
```

# What this document does not verify

The sweep is anchored on `brimmest()`, so it can only reach claims `brimmest()`
can express: whether a reported (mean, SD) is attainable by integer data on a
bounded scale, under a given scoring and precision. Four things fall outside
that, and are checked in `validate_sd_bounds_functions.qmd` instead.

**The alpha-conditional bounds.** `brimmest()` has no `alpha` argument — the
attainable lattice it builds is over sum and sum of squares, which carry no
information about how a composite decomposes into items. So nothing here tests
the internal-consistency ceiling, the amplified floor, or the sharp Gini
envelope of Theorem H5. Those need their own brute-force allocation searches.

**Sharpness.** A sweep can show the bounds are never violated. It cannot show
they are *attained*, which is the stronger claim that they cannot be tightened.
That requires exhibiting the enumerated configuration that sits exactly on each
bound.

**The POMP transforms and derived builders.** `pomp_sd_parity` and
`pomp_sd_sharp`, `brimmer_map()`, `sd_bounds_curve()` and `plot_*()` are
verified by identity rather than by oracle.

**The gap-ladder threshold**, which bounds how close to the ceiling a blind-spot
cell can hide and is what makes selective reconstruction affordable.

So `validate_sd_bounds_functions.qmd` is not redundant with this file. The
division is: that document verifies *claims about functions*, this one verifies
*verdicts over grids*.

# Assertions

```{r assertions}
tibble(check = names(checks), passed = unlist(checks)) |>
  kable(caption = "Every hard claim in this document.")

stopifnot(all(unlist(checks)))
```

All assertions passed.

# Session info

```{r session}
sessionInfo()
```
