---
title: "POMP TIDES"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

library(tidyverse)
library(scales)
library(janitor)
#library(rsprite2)
library(scrutiny) 
library(tides)
library(purrr)
library(patchwork)
library(knitr)
library(kableExtra)

```

# GRIM vs GRIM+GRIMMER vs GRIM+GRIMMER+Umbrella plot 

Using {scrutiny} + {tides}

```{r}

# dat_n_11_tides <-
#   expand_grid(mean = seq(from = 0.5, to = 7.5, by = 0.01),
#               sd = seq(from = 0, to = 4, by = 0.01)) |>
#   mutate(n_obs   = 11,
#          m_prec  = 2,
#          sd_prec = 2,
#          n_items = 1,
#          min_val = 1,
#          max_val = 7,
#          calculate_min_sd = TRUE) |>
#   mutate(grim = pmap(list(as.character(mean), n_obs),
#                      grim)) |>
#   mutate(grimmer = pmap(list(as.character(mean), as.character(sd), n_obs),
#                         grimmer)) |>
#   mutate(tides = pmap(list(mean, sd, n_obs, min_val, max_val, n_items, sd_prec,
#                            calculate_min_sd = TRUE, verbose = FALSE),
#                       tides_single)) |>
#   unnest(grim) |>
#   unnest(grimmer) |>
#   unnest(tides)
# 
# dat_n_100_tides <-
#   expand_grid(mean = seq(from = 0.5, to = 7.5, by = 0.01),
#               sd = seq(from = 0, to = 4, by = 0.01)) |>
#   mutate(n_obs   = 100,
#          m_prec  = 2,
#          sd_prec = 2,
#          n_items = 1,
#          min_val = 1,
#          max_val = 7,
#          calculate_min_sd = TRUE) |>
#   mutate(grim = pmap(list(as.character(mean), n_obs),
#                      grim)) |>
#   mutate(grimmer = pmap(list(as.character(mean), as.character(sd), n_obs),
#                         grimmer)) |>
#   mutate(tides = pmap(list(mean, sd, n_obs, min_val, max_val, n_items, sd_prec,
#                            calculate_min_sd = TRUE, verbose = FALSE),
#                       tides_single)) |>
#   unnest(grim) |>
#   unnest(grimmer) |>
#   unnest(tides)
#
# dat_n_5_to_100_tides <-
#   expand_grid(mean = seq(from = 0.5, to = 7.5, by = 0.01),
#               sd = seq(from = 0, to = 4, by = 0.01),
#               n_obs = seq(from = 5, to = 100, by = 5)) |>
#   mutate(m_prec  = 2,
#          sd_prec = 2,
#          n_items = 1,
#          min_val = 1,
#          max_val = 7,
#          calculate_min_sd = TRUE) |>
#   mutate(grim = pmap(list(as.character(mean), n_obs),
#                      grim)) |>
#   mutate(grimmer = pmap(list(as.character(mean), as.character(sd), n_obs),
#                         grimmer)) |>
#   mutate(tides = pmap(list(mean, sd, n_obs, min_val, max_val, n_items, sd_prec,
#                            calculate_min_sd = TRUE, verbose = FALSE),
#                       tides_single)) |>
#   unnest(grim) |>
#   unnest(grimmer) |>
#   unnest(tides)
# 
# write_rds(dat_n_11_tides, "results_grim_grimmer_tides_n_11_tides.rds")

dat_n_11_tides <- read_rds("results_grim_grimmer_tides_n_11_tides.rds")

```


## TIDES plot

```{r}

# doesn't work for so many points
# dat_n_11_tides |>
#   rename(n = n_obs,
#          min = min_val,
#          max = max_val,
#          digits = sd_prec) |>
#   plot_tides_multiple()

set.seed(42)

dat_n_11_tides |>
  sample_n(size = 30) |>
  rename(n = n_obs,
         min = min_val,
         max = max_val,
         digits = sd_prec) |>
  plot_tides_multiple()


set.seed(42)

dat_n_11_tides |>
  sample_n(size = 1) |>
  rename(n = n_obs,
         min = min_val,
         max = max_val,
         digits = sd_prec) |>
  plot_tides_single(text_size = 0.6)

```

## POMP plot

The big issue with the pomp plot is that it only plots values whose SD range is calculable, so many values that would be inside the bounds but are GRIM/GRIMMER/TIDES inconsistent are not even plotted. How to calculate bounds in a more abstract way and plot all values inside it?

Why is the point just inside the rectangle inconsistent when it should be consistent?

### dev wrangling

```{r}

dat_liberal_boundaries_n_11 <- dat_n_11_tides |>
  # Make two copies of 'min_sd'
  mutate(min_sd_up = min_sd,
         min_sd_down = min_sd,
         max_sd_up = max_sd,
         max_sd_down = max_sd) |>
  # Fill upwards
  fill(min_sd_up, .direction = "up") |>
  fill(max_sd_up, .direction = "up") |>
  # Fill downwards
  fill(min_sd_down, .direction = "down") |>
  fill(max_sd_down, .direction = "down") |>
  # Take the minimum non-NA value between the two
  mutate(min_min_sd = pmin(min_sd_up, min_sd_down, na.rm = TRUE),
         min_min_sd = case_when(mean >= min_val & mean <= max_val ~ min_min_sd,
                                TRUE ~ NA),
         max_max_sd = pmax(max_sd_up, max_sd_down, na.rm = TRUE),
         max_max_sd = case_when(mean >= min_val & mean <= max_val ~ max_max_sd,
                                TRUE ~ NA)) |>
  # Remove helper columns
  select(-min_sd_up, -min_sd_down, -max_sd_up, -max_sd_down) |>
  # Find limits - min and max values of min and max sd
  group_by(mean, 
           min_val, max_val, n_obs, n_items,
           min_sd, max_sd) |>
  summarize(min_sd = min(min_min_sd),
            max_sd = max(max_max_sd),
            .groups = "drop") |>
  drop_na()

ggplot(dat_liberal_boundaries_n_11) +
  geom_ribbon(aes(x = mean, ymin = min_sd, ymax = max_sd), fill = "grey30", alpha = 0.25) +
  theme_linedraw() +
  scale_x_continuous(breaks = breaks_pretty(n = 7),
                     name = "Mean") +
  scale_y_continuous(breaks = breaks_pretty(n = 7),
                     name = "Standard Deviation") +
  ggtitle("Liberal boundary of min and max SD for N=11\nIgnoring granular consistency of mean and SD")

```

- This is the generic solution i needed, i think?
- but then how did the psymeta code calculate SD bounds for all values?

# ?

```{r}

dat_liberal_boundaries_n_11 |>
  mutate(pomp_mean = case_when(mean >= min_val & mean <= max_val ~ 
                                 round_half_up((mean - min_val)/(max_val - min_val), 2),
                               TRUE ~ NA),
         pomp_sd = case_when(!is.na(min_sd) & !is.na(max_sd) ~
                               round_half_up((sd - min_sd)/(max_sd - min_sd), 2),
                             TRUE ~ NA),
         pomp_sd = case_when(is.infinite(pomp_sd) ~ 0,
                             TRUE ~ pomp_sd)) |>
  ggplot() +
  geom_point(aes(x = pomp_mean, y = pomp_sd)) +
  theme_linedraw() +
  scale_x_continuous(breaks = breaks_pretty(n = 7),
                     name = "Mean") +
  scale_y_continuous(breaks = breaks_pretty(n = 7),
                     name = "Standard Deviation") +
  ggtitle("Liberal boundary of min and max SD\nIgnoring granular consistency of mean and SD")


dat_liberal_boundaries_n_11 |>
  select(pomp_mean, pomp_sd) |> #tides_consistent
  drop_na() |>
  distinct(pomp_mean, pomp_sd, .keep_all = TRUE) |>
  #arrange(tides_consistent) |>
  ggplot(aes(pomp_mean, pomp_sd)) + #, color = tides_consistent
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 1, ymax = Inf), fill = "grey91", color = NA) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = -Inf, ymax = 0), fill = "grey91", color = NA) +
  geom_rect(aes(xmin = -Inf, xmax = 0, ymin = -Inf, ymax = Inf), fill = "grey91", color = NA) +
  geom_rect(aes(xmin = 1, xmax = Inf, ymin = -Inf, ymax = Inf), fill = "grey91", color = NA) +
  geom_point(alpha = 0.7, shape = 15, size = 1) + 
  scale_x_continuous(breaks = c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1),
                     labels = scales::label_percent(),
                     name = "Proportion-of-Maximum-Possible Mean") +
  scale_y_continuous(breaks = function(x) unique(c(scales::breaks_pretty(n = 10)(x), -1, 0, 1)),
                     labels = scales::label_percent(),
                     name = "Proportion-of-Maximum-Possible SD",
                     trans = signed_log10_trans) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank()) + 
  scale_color_viridis_d(begin = 0.3, end = 0.7) +
  labs(color = "Inside possible ranges") +
  guides(color = guide_legend(override.aes = list(size = 4)))

```

### Plot assuming pomp values returned from tides_single() are sufficient

note that they only contain 

```{r}

plot_tides_multiple_pomp <- function(dat, shade_limits = "inside"){
  
  signed_log10_trans <- trans_new(
    name = "signed_log10",
    transform = function(x) sign(x) * log10(abs(x) + 1),
    inverse = function(x) sign(x) * (10^abs(x) - 1)
  )
  
  if(shade_limits == "outside"){
    
    p <- dat |>
      select(pomp_mean, pomp_sd, tides_consistent) |>
      drop_na() |>
      distinct(pomp_mean, pomp_sd, .keep_all = TRUE) |>
      arrange(tides_consistent) |>
      ggplot(aes(pomp_mean, pomp_sd, color = tides_consistent)) +
      geom_rect(aes(xmin = 0, xmax = 1, ymin = 1, ymax = Inf), fill = "grey91", color = NA) +
      geom_rect(aes(xmin = 0, xmax = 1, ymin = -Inf, ymax = 0), fill = "grey91", color = NA) +
      geom_rect(aes(xmin = -Inf, xmax = 0, ymin = -Inf, ymax = Inf), fill = "grey91", color = NA) +
      geom_rect(aes(xmin = 1, xmax = Inf, ymin = -Inf, ymax = Inf), fill = "grey91", color = NA) +
      geom_point(alpha = 0.7, shape = 15, size = 1) + 
      scale_x_continuous(breaks = c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1),
                         labels = scales::label_percent(),
                         name = "Proportion-of-Maximum-Possible Mean") +
      scale_y_continuous(breaks = function(x) unique(c(scales::breaks_pretty(n = 10)(x), -1, 0, 1)),
                         labels = scales::label_percent(),
                         name = "Proportion-of-Maximum-Possible SD",
                         trans = signed_log10_trans) +
      theme_linedraw() +
      theme(panel.grid.minor = element_blank()) + 
      scale_color_viridis_d(begin = 0.3, end = 0.7) +
      labs(color = "Inside possible ranges") +
      guides(color = guide_legend(override.aes = list(size = 4)))
    
  } else if (shade_limits == "inside") {
    
    p <- dat |>
      select(pomp_mean, pomp_sd, tides_consistent) |>
      drop_na() |>
      distinct(pomp_mean, pomp_sd, .keep_all = TRUE) |>
      arrange(tides_consistent) |>
      ggplot(aes(pomp_mean, pomp_sd, color = tides_consistent)) +
      geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1), fill = "grey91", color = NA) +
      geom_point(alpha = 1, shape = 15, size = 1.5) + # ,  , 
      scale_x_continuous(breaks = c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1),
                         labels = scales::label_percent(),
                         name = "Proportion-of-Maximum-Possible Mean") +
      scale_y_continuous(breaks = function(x) unique(c(scales::breaks_pretty(n = 8)(x), -1, 0, 1)),
                         labels = scales::label_percent(),
                         name = "Proportion-of-Maximum-Possible SD",
                         trans = signed_log10_trans) +
      theme_linedraw() +
      theme(legend.position = "top",
            legend.direction = "vertical",
            panel.grid.minor = element_blank()) + 
      scale_color_manual(values = c("TRUE" = "#43BF71FF", 
                                    "FALSE" = "#35608DFF"),
                         labels = c("TRUE" = "TIDES-consistent value", 
                                    "FALSE" = "TIDES-inconsistent value")) +
      labs(color = "") +
      guides(color = guide_legend(override.aes = list(size = 4)))
    
  } else {
    stop("shade_limits must be 'inside' or 'outside")
  }
  
  return(p)
}

```

```{r fig.height=6, fig.width=7}

plot_tides_multiple_pomp(dat_n_11_tides)

```

#### ?

```{r}

# could 
# dat_n_11_tides |>
#   fill(pomp_sd, .direction = "up") |>
#   sample_n(size = 30) |>
#   plot_tides_multiple_pomp()

temp <- dat_n_11_tides |>
  mutate(pomp_mean = janitor::round_half_up(pomp_mean, 2), ## 
         pomp_sd = janitor::round_half_up(pomp_sd, 2)) |>
  group_by(pomp_mean) |>
  filter(abs(pomp_sd) == max(abs(pomp_sd), na.rm = TRUE)) |>
  ungroup() |>
  distinct(pomp_mean, pomp_sd, .keep_all = TRUE)

plot_tides_multiple_pomp(temp)


#plot_tides_multiple_pomp(dat_n_11_tides, shade_limits = "outside")

```

#### Random subset

```{r fig.height=6, fig.width=7}

set.seed(42)

dat_n_11_tides |>
  filter(sd_range_calculable == TRUE) |>
  sample_n(size = 30) |>
  plot_tides_multiple_pomp()

```




