---
title: "The BRIMMEST test: exact certification of reported means and standard deviations on bounded scales"
author:
  - name: Ian Hussey
    email: ian.hussey@unibe.ch
    affiliations:
      - name: University of Bern
abstract: 'Error-detection tests for reported summary statistics, including GRIM, GRIMMER, and the standard-deviation bounds tests, are necessary conditions: failing one proves a reported (mean, SD, n) is impossible, but passing all of them proves nothing. A residual set of reports clears every such screen and still corresponds to no possible dataset. We introduce BRIMMEST, an exact certification test for reported means and standard deviations on bounded integer scales. It enumerates the complete attainable (mean, SD) lattice for a design by dynamic programming over reachable (sum, sum of squares) states, rounds that lattice to the reporting precision, and tests membership. A hit proves that some real sample produces exactly the reported values; a miss proves that none does, relative to the admitted rounding conventions. The verdict is therefore exact in both directions, which no test in the GRIM family achieves. BRIMMEST agrees with brute-force enumeration and with the CLOSURE reconstruction algorithm on every case tested (zero disagreements across roughly 5,700 grid cells), while running two to three orders of magnitude faster than reconstruction, because its cost depends only on the scale limits and sample size rather than on how many datasets exist. Sweeping 556,975 reportable tuples across 13 designs, we quantify the blind spot of the closed-form screens and show it is not monotone in sample size. We also report implementation errors that the exact oracle exposed in widely used software, in both error directions. We situate BRIMMEST relative to CLOSURE and SPRITE, and recommend a three-stage workflow: closed-form screens first, BRIMMEST to settle survivors, and reconstruction only when the candidate datasets themselves are needed. The test is implemented as brimmest() in the strait R package.'
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

<!-- TODO before submission:
  - confirm the acronym expansion (docstring says "Bounds-Related Inconsistency
    of Means and Errors, Settled Test"; BRIMMER expands with "SDs" not "Errors")
  - file the multi-item scrutiny issue so section 8 can cite it
  - complete placeholder references marked [REF]
  - decide venue; this draft is written long, cut to fit -->

```{r include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_unsum <- requireNamespace("unsum", quietly = TRUE)
library(strait)
library(ggplot2)
```

# Introduction: necessary is not sufficient

A reported mean, standard deviation, and sample size are not free to take any
values. When the underlying data are integers on a bounded scale, as with
Likert-type items, counts, and most questionnaire measures, the reported
statistics must be producible by some sample of $n$ integers between the scale
limits, after rounding to the reported precision. Tests that exploit this fact
have become standard tools of error detection. GRIM checks that the reported
mean is a multiple of $1/n$ at the reported precision (Brown & Heathers, 2017).
GRIMMER extends the logic to the standard deviation (Anaya, 2016). The
SD-bounds tests check the reported SD against closed-form minimum and maximum
bounds given the scale limits, $n$, and the mean (Hussey et al., in prep).

Each of these is a necessary condition. Failing one proves the report
impossible. Passing all of them proves nothing, because none of them asks
whether the mean and SD are *jointly* attainable by a single sample. In
practice, a report that clears every screen is routinely described as
"consistent", a word that quietly overstates what was shown.

The gap is not hypothetical. On a 1--5 scale with $n = 9$, reporting to one
decimal place, the pair mean $= 1.3$, SD $= 0.9$ lies inside the SD bounds,
passes GRIM, and passes GRIMMER. No sample of nine integers between 1 and 5
produces it:

```{r}
brimmest(l = 1, u = 5, n = 9, mean = 1.3, sd = 0.9, digits_mean = 1, digits_sd = 1)
```

This paper introduces the test that settles such cases. BRIMMEST (Bounds-Related
Inconsistency of Means and Errors, Settled Test) is an exact
certification procedure for reported (mean, SD) pairs on bounded integer
scales, implemented as `brimmest()` in the strait R package. It enumerates the
complete attainable (mean, SD) lattice for the design by dynamic programming,
rounds the lattice to the reporting precision, and tests whether the reported
pair is a member. A hit proves possibility: some sample rounds to exactly this
report. A miss proves impossibility, relative to the rounding conventions
admitted. The verdict is exact in both directions, which is precisely what the
GRIM family cannot offer.

The name completes a series. GRIM and GRIMMER test granularity; brim and
brimmer, their bounds-based counterparts in strait, test magnitude. All four
are necessary conditions of increasing strength. BRIMMEST is superlative in
the grammatical sense and, within its scope, in the technical one: no stronger
test of the same question exists, because the question is decided. It is worth
being clear at the outset that this makes it a different kind of test rather
than a stronger member of the same family. The others are closed-form
inequalities evaluated in constant time; BRIMMEST is a decision procedure with
a computational cost, albeit a modest one.

The question BRIMMEST answers is the one the CLOSURE algorithm answers by
reconstruction (Jung, [REF]): does any integer dataset match this report?
CLOSURE searches the sample space and returns the datasets it finds. BRIMMEST
constructs the reachable set of summary statistics directly and returns only
the verdict. One might call it an analytic CLOSURE certification: the same
certificate, obtained by construction rather than search. The trade, examined
in section 6, is that no witness datasets are produced, in exchange for a cost
that depends only on the scale limits and $n$, not on how many datasets happen
to exist, and for the ability to certify every report of a design against one
lattice built once.

The paper proceeds as follows. Section 2 measures the blind spot of the
closed-form screens, since the case for an exact test rests on the residual it
removes. Section 3 presents the algorithm. Section 4 reports its verification
against brute-force enumeration and against CLOSURE. Section 5 develops the
asymmetry between the two directions of the certificate, which governs how the
test must be used in practice. Section 6 positions BRIMMEST against CLOSURE
and SPRITE. Sections 7 to 9 cover practical usage, implementation errors the
exact oracle exposed in existing software, and the limits of the method's
scope. Section 10 concludes.

# How big is the blind spot?

If the closed-form screens missed nothing, an exact test would be redundant.
We measured the residual two ways.

## Grid-wise: every reportable tuple of a design

For a given design (scale limits, $n$, reported decimal places), the
reportable (mean, SD) pairs form a finite grid, the umbrella of feasible cells
under the SD bounds. `umbrella_data()` applies the full closed-form screen to
every cell: the mean must lie in $[l, u]$ and pass rounding-aware GRIM, and
the SD's rounding interval must intersect the closed-form bounds and pass
GRIMMER. Certifying the same cells with `brimmest()` then splits the
screen-passers into the truly possible and the blind spot.

```{r}
#| label: fig-blindspot
#| fig-cap: 'The blind spot of the closed-form screens. Every reportable (mean, SD) cell for a 1--5 scale, n = 9, one decimal place. Grey cells fail at least one closed-form test (bounds, GRIM, or GRIMMER). Blue cells pass the screens and are certified possible. Red cells pass every closed-form screen and are certified impossible: no sample of nine integers on 1--5 rounds to them. The annotated cell is the example from the introduction.'
#| fig-width: 7
#| fig-height: 4.5
um <- umbrella_data(n = 9, l = 1, u = 5, digits = 1, granularity = "integer")
cert <- brimmest(l = 1, u = 5, n = 9, mean = um$mean, sd = um$sd, digits_mean = 1, digits_sd = 1)

um$status <- ifelse(!um$consistent, "fails a screen",
             ifelse(cert$possible, "possible", "blind spot"))

ggplot(um, aes(mean, sd, fill = status)) +
  geom_tile(width = 0.1, height = 0.1, colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = c("fails a screen" = "grey85",
                               "possible" = "#1d4ed8",
                               "blind spot" = "#dc2626")) +
  annotate("segment", x = 1.85, y = 1.45, xend = 1.36, yend = 0.97,
           arrow = arrow(length = unit(0.15, "cm"))) +
  annotate("text", x = 1.9, y = 1.52, hjust = 0, size = 3.2,
           label = "mean 1.3, sd 0.9: passes every screen,\nno such sample exists") +
  labs(x = "reported mean", y = "reported SD", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
c(cells            = nrow(um),
  passes_screen    = sum(um$consistent),
  certified        = sum(cert$possible),
  blind_spot       = sum(um$consistent & !cert$possible),
  wrongly_screened = sum(!um$consistent & cert$possible))
```

Two numbers matter. The blind spot count is the price of using closed forms:
reports the screen lets through that are nonetheless impossible. The
`wrongly_screened` count is zero, here and everywhere we have looked: the
closed-form screens never reject a report that real data can produce. The
screens are safe; they are merely incomplete.

At scale, the same comparison was run across 13 designs spanning 1--5, 1--7,
and 0--6 scales, $n$ from 8 to 30, at one and two reported decimal places:
556,975 reportable tuples in all. The bounds produced no false flags in any of
them, and 758 blind-spot cells. The full sweep, with per-design tables and the
assertion framework that guards every number, is reported in the validation
supplement (`validate_brimmer_against_brimmest.qmd` in the strait repository);
this article does not re-run it.

## The blind spot is not monotone in n

It is tempting to assume the residual vanishes with sample size. It does not,
and it does not even shrink monotonically. On a 1--5 scale at two decimal
places, the share of screen-passing cells that certify impossible was measured
at 9.0% ($n = 10$), 4.8% ($n = 20$), 6.7% ($n = 40$), and 1.8% ($n = 80$).

```{r}
#| label: fig-prevalence
#| fig-cap: 'Blind-spot share of screen-passing cells, 1--5 scale, two decimal places, as measured in the validation supplement against the scrutiny build current at the time (see section 8). The non-monotonicity is real: the spike at n = 40 is the fingerprint of a floating-point boundary error in the screening software (scrutiny issue #86), which produces additional GRIMMER errors when n is a multiple of 40 at two decimal places. Larger samples do not obviate exact certification.'
#| fig-width: 5.5
#| fig-height: 3
prev <- data.frame(n = factor(c(10, 20, 40, 80)),
                   pct = c(9.0, 4.8, 6.7, 1.8))
ggplot(prev, aes(n, pct)) +
  geom_col(fill = c("grey35", "grey35", "#dc2626", "grey35"), width = 0.6) +
  labs(x = "sample size", y = "blind spot, % of screen-passing cells") +
  theme_minimal()
```

Two forces are at work. The genuine, method-level blind spot shrinks as the
lattice of attainable pairs densifies with $n$. Implementation errors in the
screening software do not shrink with it, and at particular sample sizes they
dominate (section 8). Both are reasons to certify rather than assume.

## Within GRIMMER's band: the missing achievability check

A second measurement isolates the method-level gap in GRIMMER itself, as
distinct from the software that implements it. Within the band of SDs that
GRIMMER can discriminate at all, the share of GRIMMER-consistent verdicts that
are provably impossible grows from roughly 4% at $n = 10$ to roughly 97% at
$n = 1000$, and does so identically in two independent implementations
(scrutiny and rsprite2), so the gap belongs to the published method rather
than to either codebase.

The gap has a closed form. Writing the integer sample sum as $s = nm + r$ with
$0 \le r < n$, an integer sample with sum $s$ has sum of squares at least

$$
q_{\min}(s, n) = n m^2 + 2 m r + r,
$$

attained by the two-point sample with $n - r$ observations at $m$ and $r$ at
$m + 1$. Equivalently, the sample SD has the granularity floor
$s_{\min} = \sqrt{r(n-r)/(n(n-1))}$. GRIMMER as published tests the parity and
granularity of the reconstructed sum of squares but never checks $q \ge
q_{\min}(s, n)$, so it accepts sums of squares below the attainable floor.
Adding the check makes GRIMMER exact for $n \ge 5$; samples of 2 to 4 retain
gaps of the sums-of-few-squares kind that no closed form repairs. We note this
repair because it is cheap and scale-free, and a repaired GRIMMER is a better
first-line screen. It does not remove the need for certification: the repaired
test is still necessary-only for $n < 5$, and it still checks the SD against
the mean's sum only through the reconstructed $q$, not against joint
attainability on a bounded scale.

# The method

## The attainable lattice

Fix integer scale limits $l \le u$, sample size $n$, and (for multi-item
composites) the number of items $k$. Shift and scale the data to
$y_i = m_g (x_i - l) \in \{0, 1, \dots, W\}$, where $W = m_g (u - l)$ and
$m_g = k$ for mean-scored composites and $1$ otherwise. Every sample statistic
we need is a function of the pair

$$
(S, Q) = \Big( \sum_i y_i, \; \sum_i y_i^2 \Big),
$$

with mean $= l + S / (n m_g)$ and
SD $= \sqrt{(Q - S^2/n) / (n-1)} \, / \, m_g$. Two samples with the same
$(S, Q)$ are indistinguishable to a (mean, SD) report, so the object to
enumerate is not the sample space but the set of reachable $(S, Q)$ states.

That set obeys a recurrence: the states reachable with $t$ observations are
the states reachable with $t - 1$, each extended by one more value
$y \in \{0, \dots, W\}$, which adds $y$ to $S$ and $y^2$ to $Q$. Dynamic
programming over this recurrence, $n$ steps of a boolean table, enumerates the
exact attainable lattice. No sample is ever constructed and no distributional
information survives; the lattice records which pairs are reachable, not how.

## Making it fast

The naive table over $(S, Q)$ has $nW \times nW^2$ cells and is wasteful in
ways that compound. The strait implementation applies five reductions, all in
plain R, which together speed the enumeration by a factor of 23 to 29:

1. **A better second axis.** Replace $Q$ by $R = WS - Q = \sum_i y_i (W - y_i)$.
   Each observation contributes at most $\lfloor W^2/4 \rfloor$ to $R$, against
   $W^2$ to $Q$, so the axis is four times shorter, and $R \ge 0$ always.
2. **Common divisor.** The per-value contributions $y(W - y)$ share a greatest
   common divisor $g$ ($g = 2$ whenever $W$ is odd); dividing the axis by $g$
   halves it again in the common case.
3. **Growing window.** After $t$ observations only $tW + 1$ values of $S$ are
   reachable, so early steps operate on a small corner of the table rather
   than all of it.
4. **Reflection.** The substitution $y \to W - y$ maps $S \to tW - S$ and
   leaves $R$ fixed, so the table is symmetric; compute the lower half of the
   $S$ axis and mirror it.
5. **Byte storage.** One-byte `raw` cells rather than four-byte logicals keep
   the working set in cache.

One implementation lesson is worth recording because it inverts the usual
advice. Dense whole-table operations were six times slower than a sparse
scatter of reachable states, because reachable cells are only about 7% of the
naive grid; vectorised block operations only begin to win after reductions 1
to 4 shrink the table around the occupied region. Measured end to end, a 1--7
scale at $n = 61$ fell from 2.57 s to 0.098 s. The lattice reaches $n = 300$
on a 0--6 scale (1.57 million attainable pairs) in about 11 s and $n = 400$ on
1--7 (2.8 million pairs) in about 23 s, well past the point where
reconstruction-based methods are uncomfortable. A guard (`max_cells`, default
$2 \times 10^7$) makes the enumeration fail cleanly rather than exhaust memory
on designs beyond its reach, such as wide mean-scored composites at large $n$.

## Certifying one report without enumerating every other

The five reductions shrink the table but leave the work proportional to the
whole design. For a single report that is a membership question answered by a
reachability-of-everything computation, and past a certain scale width the
guard refuses outright: a 0--63 inventory at $n = 50$ would need a
$3151 \times 24801$ table. Yet a report pins both coordinates almost
completely.

The mean's rounding interval admits only integer sums
$S \in [n(m_{\text{lo}} - l),\, n(m_{\text{hi}} - l)]$ — typically one to five
values rather than thousands. This is the GRIM condition, read as a constraint
that generates candidates rather than as a test that rejects. Each candidate
$S$ then pins the sum of squares to a narrow window, because
$Q = S^2/n + (n-1)\,s^2$ and $s$ is known to its reporting precision. The
question becomes: can $n$ integers in $[0, W]$ realise one of a handful of
$(S, Q\text{-window})$ targets? Three consequences follow.

**Inverse rounding.** The reported values are unrounded to intervals once,
rather than the lattice being forward-rounded cell by cell under every admitted
rule.

**A corridor rather than a rectangle.** A partial sum $s$ after $t$
observations is viable only if the remaining $k = n - t$ can still carry it to
some target, giving
$s \in [S^\ast - kW,\, \min(tW, S^\ast)]$, and likewise
$R \in [R^\ast_{\min} - k\lfloor W^2/4\rfloor / g,\, R^\ast_{\max}]$. States
outside are never allocated, let alone visited.

**Early exit in one direction.** An observation with $y = 0$ contributes
nothing to either coordinate, so a target reached after $t < n$ observations
pads out to exactly $n$ with scores at the scale minimum. Reachability is
therefore monotone in $t$, and a hit can be declared the moment it appears.
Possibility is often immediate; impossibility still requires sweeping the whole
corridor. The asymmetry runs in the useful direction for screening, where most
reports are real.

Measured, the corridor holds about 9% of the full table's cells across the
designs tested. The wall-clock gain is smaller than that ratio suggests —
roughly a factor of two on large designs — because the per-layer windows
forfeit the reflection symmetry of reduction 4 and add bookkeeping the dense
sweep does not pay. The decisive gain is categorical rather than
proportional: designs the lattice refuses become answerable at all. The 0--63
inventory at $n = 50$ certifies in about 105 s where the lattice route cannot
run, and a 0--20 scale at $n = 30$ falls from 0.97 s to 0.44 s. Small designs
keep the lattice route, which is cheaper there and is cached per
$(l, u, n)$, so screening a table of reports on one design pays for the
enumeration once.

The two routes are interchangeable by construction, and are verified to be so:
their verdicts agree over 1.75 million reporting-grid cells spanning six
designs, two reporting precisions and two rounding rules.

## Replacing the sweep with arithmetic and a witness

The corridor still sweeps. Its cost is set by the design, not by the report,
which is what makes the scale width a barrier rather than an inconvenience.
Two observations remove the sweep for almost every report.

**The sandwich.** At a fixed sum $S$, the achievable sums of squares lie
between the clustered configuration and the Structure S one — the integer
forms of the floor and ceiling the closed-form screens already compute:

$$
q_{\min}(S) = (n - r)m^2 + r(m+1)^2, \quad m = \lfloor S/n \rfloor,\ r = S - nm
$$
$$
q_{\max}(S) = \nu W^2 + \rho^2, \quad \nu = \lfloor S/W \rfloor,\ \rho = S - \nu W
$$

and every sample satisfies $Q \equiv S \pmod 2$, because $y^2 \equiv y
\pmod 2$ term by term. A window that meets no integer of the sum's parity
inside $[q_{\min}, q_{\max}]$ is impossible on arithmetic alone, in $O(1)$ per
candidate sum. The parity condition is not a refinement of the bounds: it
rejects reports comfortably inside them, and it does so at a rate that does
not vanish with $n$. On a 0--100 scale at $n = 100$, a mean of 99.80 with an
SD of 0.50 is disposed of by parity, the sole integer its SD window admits
being odd where the sum is even.

**The witness.** What survives the sandwich goes to a constructive search over
the sample itself, ordered non-increasingly, which is to say over the
partitions of $S$ into at most $n$ parts of size at most $W$. At every step
the sandwich is re-applied to what remains to be placed, with the cap set by
the value just chosen, and the children are tried in order of how centrally
the surviving window sits inside each child's own sandwich. Reaching a
complete sample proves possibility and hands back a witness; exhausting the
tree proves impossibility.

The search is depth-first on an explicit stack rather than recursive, so its
depth — one level per observation placed — is not a constraint: a report at
$n = 20000$ is certified, with a witness, in under half a second.

Two structural choices keep the tree small. Requiring the sequence to be
non-increasing collapses the $n!$ orderings of a sample to one. Reflecting
$y \mapsto W - y$ whenever $S$ sits above $nW/2$ — under which $S \mapsto
nW - S$ and $Q \mapsto Q + nW^2 - 2WS$ — means the search always builds from
the nearer wall, where the partition tree is shallow. Near either wall the
surviving tree is a handful of nodes, which is precisely where the closed-form
screens leak: the blind-spot cell of the introduction has one candidate sum
and three reachable values of $Q$, so the certificate costs three
configurations rather than a corridor sweep.

This subsumes the enumeration of the ceiling's top rungs as a special case,
and it is strictly more informative than a greedy construction would be: a
greedy walk upward from the clustered configuration certifies possibility when
it succeeds and concludes nothing when it fails, whereas an exhausted search
is a proof in both directions.

Measured against the corridor it replaces, on medians of repeated runs
covering both rounding rules, the blind-spot cell falls from 493 to 41
microseconds and the 0--63 inventory at $n = 50$ from 78 s to 0.63 ms. A
0--100 scale at $n = 100$ — untimed before, because neither sweeping route
would run — certifies in 1.5 ms, and the parity screen alone disposes of a
report there in 25 microseconds. A node of the search costs about 6
microseconds, of which the arithmetic is a small part: the per-node cost is
flat from $W = 20$ to $W = 100$, so what is being paid for is interpretation,
not computation. A compiled kernel would be the obvious next step, and would
change these figures by roughly two orders of magnitude, but it would not
change what is decidable. Over 2.72 million reporting-grid cells
spanning six designs, two precisions and two rounding rules, the search agreed
with the lattice route on every cell and was never once cut short by its node
budget. Grid-sized workloads keep the cached lattice route, which remains far
cheaper per cell when the whole design is being enumerated anyway; the search
is for the case it was built for, a report at a time.

What remains swept is a sliver: reports whose search neither completes nor
exhausts within budget, for which the corridor still answers. No such report
appeared in the validation sweep, but the fallback is retained because its
absence there is a measurement, not a theorem.

## An open problem: the near-closed-form certificate

Both routes still sweep a state space whose size is set by the design rather
than by the report, which is why a guard is needed at all. The blind-spot
geometry suggests this may be removable.

At fixed $S$, the attainable values of $Q$ appear to be contiguous — every
integer between the floor and some threshold is realised — with gaps confined
to a fringe hugging the mean-conditional ceiling. Across every design examined
here, all blind-spot cells lay in the top fringe of the umbrella's height, and
the gap structure there is computable directly by best-first search outward
from the Structure S maximum configuration using inward unit transfers, which
enumerates the top rungs completely.

If contiguity below the threshold can be proven, certification reduces to
arithmetic. For a target below the threshold, membership follows from the
floor, the threshold and a congruence condition, with no enumeration at all.
For a target in the fringe, the ladder search explores only the handful of $Q$
values adjacent to it. Certification would then be near-closed-form:
independent of $n$ and of scale width, and the `max_cells` guard could be
retired. That leaves the contiguity theorem as the substantive open problem
this method poses, and the one whose resolution would matter most for applying
exact certification at the scale widths clinical instruments actually use.

## From lattice to verdict

The lattice holds exact statistics; reports are rounded. `brimmest()` rounds
every lattice pair to the reported precision under each admitted rounding rule
and tests whether the reported pair is a member of the rounded set. Matching
is done in integer grid steps (a mean reported to 2 dp becomes a whole number
of hundredths), which is exact and sidesteps floating-point comparison
entirely. The verdict is returned per report, together with the rounding rules
under which it is reachable:

```{r}
brimmest(l = 1, u = 5, n = 9, digits_mean = 1, digits_sd = 1,
         mean = c(3.0, 1.3, 2.5, 2.0),
         sd   = c(1.0, 0.9, 1.2, 1.1))
```

One lattice is built per design, cached, and reused across every report, so
certifying a table of results costs barely more than certifying one. Where the
design is large and the reports are few, the targeted route above is taken
instead and the lattice is never built; the choice is automatic and the
verdicts are identical either way.

## Why the closed forms leak

The lattice view also explains the blind spot geometrically. The closed-form
bounds trace the smooth envelope of the attainable region; the attainable
pairs are a discrete lattice strictly inside it. Near the mean-conditional
ceiling the lattice thins, and reporting-grid cells fit between the envelope
and the nearest lattice point: inside every closed-form bound, yet empty.

```{r}
#| label: fig-lattice
#| fig-cap: 'Why closed-form screens must leak. One 0--3 item, n = 7. Points are the complete attainable (mean, SD) lattice; the shaded band is the closed-form quasi-integer envelope that the bounds tests check against. Near the ceiling the lattice thins, and a reporting grid laid over this picture has cells that lie inside the band but contain no attainable point. Those cells are the blind spot.'
#| fig-width: 6
#| fig-height: 3.8
lat  <- sd_region_data(l = 0, u = 3, n = 7, rule = "attainable")
band <- sd_region_data(l = 0, u = 3, n = 7, rule = "quasi")

ggplot() +
  geom_ribbon(data = band, aes(x = mean, ymin = lo, ymax = hi),
              fill = "grey88") +
  geom_point(data = lat, aes(mean, sd), size = 0.7, colour = "#1d4ed8") +
  labs(x = "mean", y = "SD") +
  theme_minimal()
```

# Correctness

An exact test earns the adjective by verification, not assertion. The lattice
and the verdicts it produces were checked three independent ways; the full
protocol, with an assertion block that fails the render if any check fails,
constitutes the validation supplement.

**Against brute force.** On designs small enough to enumerate every integer
sample directly, the set of (sum, sum of squares) pairs produced by exhaustive
enumeration equals the DP lattice exactly. This checks the recurrence and all
five reductions with no theory shared between the two sides.

**Against CLOSURE.** For every cell of full reporting grids across six designs
(1--5 and 1--7 scales, $n$ from 9 to 20, roughly 5,700 cells), `brimmest()`'s
verdict was compared with whether `unsum::closure_generate()` finds at least
one dataset. The two methods agree on every cell, in both directions. CLOSURE
is an independent implementation of an independent algorithm in a separate
package, so agreement here is informative rather than circular.

```{r eval=has_unsum}
closure_says <- vapply(seq_len(nrow(um)), function(i) {
  r <- try(suppressWarnings(unsum::closure_generate(
    mean = sprintf("%.1f", um$mean[i]), sd = sprintf("%.1f", um$sd[i]),
    n = 9, scale_min = 1, scale_max = 5, rounding = "up_or_down")),
    silent = TRUE)
  if (inherits(r, "try-error")) FALSE else nrow(r$results) > 0
}, logical(1))

c(cells = nrow(um),
  agree = sum(closure_says == cert$possible),
  disagree = sum(closure_says != cert$possible))
```

**Internally.** The grid builder used in the sweeps (`umbrella_data()`) was
verified equivalent to the scalar screen (`brimmer()`) on every one of 300
randomly drawn cells, so the sweep's screen verdicts are the package's screen
verdicts and the blind-spot counts of section 2 are attributable.

# The direction of the proof

The two directions of the certificate are not symmetric, and the asymmetry
governs correct use.

A **hit** proves possibility outright. The lattice pair that rounds onto the
report is an exact (sum, sum of squares) state reached by the dynamic program,
so some sample of $n$ integers on $[l, u]$ has exactly those statistics and
rounds to exactly this report under the named rule. Nothing is conditional.

A **miss** proves impossibility only relative to the rounding rules admitted.
A report unreachable under one convention may be reachable under another, and
sources rarely state how they rounded. The default therefore admits both
`half_up` and `half_down`, spanning the usual ambiguity about halves. The risk
of narrowing is concrete:

```{r}
# a real sample of 12 integers on 0-6
x <- c(rep(0, 9), 1, 1, 1)
c(mean = mean(x), sd = sd(x))

brimmest(l = 0, u = 6, n = 12, mean = 0.2, sd = 0.5, digits_mean = 1, digits_sd = 1)
```

The exact mean is 0.25, which becomes "0.2" only when halves round down. A
certification run with `rounding = "half_up"` alone would declare this real
sample's report impossible. For forensic use the costly error is the false
impossibility, so the rule set should be narrowed only when the source's
convention is actually known. Note that `brimmest()` has no `"up_or_down"`
option of the kind unrounding-based tests offer: the lattice is rounded
forward, and forward rounding needs a definite direction; admitting several
rules and taking the union plays the corresponding role.

# Positioning: verdicts and witnesses

Three families of tools now answer overlapping questions about a reported
(mean, SD, $n$) on a bounded scale.

| | closed-form screens (GRIM, GRIMMER, brim, brimmer) | BRIMMEST | CLOSURE (unsum) | SPRITE (rsprite2) |
|---|---|---|---|---|
| verdict | necessary only | exact | exact | heuristic |
| witness datasets | none | none | all | some |
| cost depends on | nothing (closed form) | $l, u, n$ only | number of matching datasets | samples requested |
| whole reporting grid | instant | one lattice, every cell | per cell | per cell |

CLOSURE reconstructs every dataset compatible with the report. That is a
strictly richer output: it gives existence, count, and the distributional
shape of the candidates, which matters when the question is what the data
could have looked like. The cost is that the search grows with the number of
solutions and with $n$ and scale width. SPRITE samples candidate datasets
heuristically; it is fast and illustrative but proves nothing in either
direction. BRIMMEST returns the verdict alone, by construction: the lattice
records reachability, not composition.

When only the verdict is needed, the difference in cost is large. Measured on
the full umbrella grid of each design (macOS, R 4.5.2):

| $n$ | scale | cells | `brimmest()` | CLOSURE | speed-up |
|----:|:------|------:|-------------:|--------:|---------:|
|   9 | 1--5  |   610 |      0.003 s |  1.80 s |    ~700x |
|  12 | 1--5  |   684 |      0.004 s |  1.72 s |    ~430x |
|  15 | 1--5  |   693 |      0.002 s |  1.62 s |    ~840x |
|  20 | 1--5  |   706 |      0.002 s |  1.72 s |    ~700x |
|  12 | 1--7  |  1497 |      0.004 s |  4.12 s |   ~1170x |
|  15 | 1--7  |  1499 |      0.007 s |  4.96 s |    ~740x |

The comparison is not adversarial, because the tools compose. The workflow we
recommend is staged by cost and by what each stage can prove:

1. **Screen** with the closed forms (GRIM, GRIMMER, brim, brimmer). Instant,
   scale-free where BRIMMEST is not, and every failure is already a proof of
   impossibility.
2. **Certify** the survivors with `brimmest()`. This settles every case the
   screens cannot, at a cost independent of how many datasets exist.
3. **Reconstruct** with CLOSURE only where the candidate datasets themselves
   are needed, for example to argue that every dataset compatible with a
   suspicious report is substantively absurd.

# Using the test

Certifying a table of reported statistics from a paper takes one call per
design. Reports sharing scale limits, $n$, and precision share a lattice, and
the function vectorises over the reports:

```{r}
brimmest(l = 1, u = 5, n = 9, digits_mean = 1, digits_sd = 1,
         mean = c(3.0, 1.3, 2.5), sd = c(1.0, 0.9, 1.2))
```

Precision can differ between the mean and SD (`digits_mean`, `digits_sd`),
and `scrutiny::decimal_places()` recovers the reported precision from the printed strings
when trailing zeros make it ambiguous.

Multi-item composites are supported through `scoring` and `n_items`. Sum
scores of $k$ items on $[a, b]$ are integers on $[ka, kb]$ and reduce to the
single-item case; mean scores live on a $1/k$ grid, which the lattice handles
by setting $m_g = k$. The distinction matters in practice: reports that are
possible on the mean-score grid can be declared impossible by screens that
mishandle the granularity of multi-item scales, an error class we return to
in section 8.

Rounding rules beyond the defaults (`native`, `ceiling`, `floor`, `trunc`,
`anti_trunc`) are available when a source's convention is known or under
audit. The `rules` column of the output names which admitted rules reach each
report, which is itself diagnostic: a report reachable only under an unusual
convention invites a closer look at the source's other numbers.

# What the exact oracle found in existing software

An exact test has a second use beyond certifying reports: it is ground truth
for the screening software itself. Sweeping whole reporting grids and
comparing screen verdicts against certification exposed three distinct errors
in scrutiny, the standard R implementation of GRIM and GRIMMER. We report
them not as criticism of a valuable package but as evidence that this
software class needs oracle-based testing, and both error directions occurred.

**Missed impossibles** (scrutiny issue #85). A conjunction bug in GRIMMER's
second and third component tests accepts some impossible reports. Witness:
mean 0.11, SD 0.87, $n = 64$ on 0--6 is impossible under `half_up`, yet
returned consistent. The same tuple is attainable under `half_down`, which is
exactly the distinction an exact, rounding-aware oracle keeps straight and a
closed-form implementation blurred.

**False flags at particular n** (scrutiny issue #86). Floating-point boundary
errors in the GRIM and GRIMMER scalar tests produce false inconsistencies at
$n$ a multiple of 40 at two decimal places (20 at one decimal place). Witness:
the sample `c(rep(1, 7), rep(0, 193))`, $n = 200$, whose true report (0.04,
0.18) under half-up is flagged. This is the error class that produced the
$n = 40$ spike in @fig-prevalence, and it is the serious direction: a false
flag is a false accusation.

**False flags on multi-item scales** (to be filed). For multi-item means,
scrutiny's `items` argument mishandles granularity. Witness: ten people each
scored as the mean of two 1--5 items, `c(2, 2, 2, 2, 2, 2, 2, 2.5, 2.5, 2.5)`,
reporting mean 2.2, SD 0.2. `brimmest()` with mean-scoring and $k = 2$
certifies it possible (it is; the data are right there); scrutiny's
`grimmer(items = 2)` flags it inconsistent under every rounding rule. The
identical data analysed on the sum scale pass, which isolates the error to
the multi-item pathway. The sweep found five such cells at $k = 2$ and two at
$k = 3$.

The general point outlives the specific bugs. Closed-form screens are
arithmetic on the boundary between possible and impossible, and boundaries
are where floating point fails. An exact combinatorial oracle has no
boundary cases of that kind, which is why grid-wise certification against it
is the right regression test for the whole software class.

# Scope

BRIMMEST is exact within a scope, and the scope has edges.

- **Requires the scale limits.** GRIM and GRIMMER are scale-free; the lattice
  is not. Where $l$ and $u$ are unknown or the measure is unbounded, the
  closed-form granularity tests remain the only options.
- **One variable at a time.** The certificate says a (mean, SD) pair is
  attainable by some sample. It says nothing about joint consistency with
  other reported statistics: a certified mean and SD may still be jointly
  impossible with a reported correlation, a subgroup breakdown, or a test
  statistic. Multivariate certification is an open problem.
- **No internal-consistency information.** The lattice is over (sum, sum of
  squares), which carries no item-level decomposition, so reported
  reliability coefficients (for example Cronbach's alpha) cannot tighten the
  verdict here. The alpha-conditional bounds of the companion article
  operate as screens on top; certifying them exactly would require
  enumeration over item allocations, a different and larger computation.
- **Integer (or 1/k-quantised) responses.** Continuous bounded data, sliders
  for example, have no lattice; there the closed-form bounds are the whole
  story.
- **Bounded designs.** The `max_cells` guard refuses designs whose state
  space exceeds the budget, such as wide mean-scored composites at large
  $n$, rather than degrading silently.
- **Misses are relative to the admitted rounding rules** (section 5). The
  default is deliberately permissive; a forensic claim of impossibility
  should quote which conventions were admitted.

Within that scope the superlative is earned: the test decides the question
the others approximate. Outside it, the necessary-condition tests are not
obsolete; they are the only tests that apply.

# Discussion

The arc from GRIM to BRIMMEST reframes what consistency-testing of reported
statistics is. The closed-form tests are best understood as fast necessary
conditions: proofs of impossibility when they fire, silence otherwise. What
has been missing is the other half, an affordable proof of possibility, and
with it the ability to say that a report's survival means something more than
that our inequalities were too loose. For bounded integer scales, the
question "could any real data produce these numbers?" is now decidable in
milliseconds per design, at any point in the reporting grid, with the same
certainty in both directions.

Three practices follow. First, reports that survive screening should be
described as certified possible or merely unscreened, not as consistent; the
word should track what was proven. Second, screening software should be
regression-tested grid-wise against an exact oracle, because the failure
modes we found (boundary arithmetic, multi-item granularity) are invisible to
spot checks and produce errors in the direction that matters most. Third,
reconstruction tools should be reserved for the cases that need
reconstruction; the common forensic question is settled more cheaply, and
exactly, without them.

The open problems are the scope edges. Joint certification of a (mean, SD)
pair with a correlation or a second group's statistics would catch
inconsistencies that no marginal test can; the state space is larger but the
same DP logic applies. Certification under reported reliability requires
enumeration over item allocations. And a repaired, scale-free GRIMMER
(section 2.3) deserves adoption independently, as the strongest test
available when scale limits are unknown. Each extension trades more
computation for more of the reported record brought under exact scrutiny,
which is the direction error detection should move: fewer heuristics, more
certificates.

# References {.unnumbered}

::: {#refs}
Anaya, J. (2016). The GRIMMER test: A method for testing the validity of
reported measures of variability. *PeerJ Preprints, 4*, e2400v1.

Brown, N. J. L., & Heathers, J. A. J. (2017). The GRIM test: A simple
technique detects numerous anomalies in the reporting of results in
psychology. *Social Psychological and Personality Science, 8*(4), 363--369.

Heathers, J. A. J., Anaya, J., van der Zee, T., & Brown, N. J. L. (2018).
Recovering data from summary statistics: Sample Parameter Reconstruction
via Iterative TEchniques (SPRITE). *PeerJ Preprints, 6*, e26968v1.

Hussey, I. (2024). strait: Bounds tests for reported summary statistics.
R package. https://github.com/ianhussey/strait

Hussey, I., Jung, L., Lakena, W., & Cummins, J. (in prep). The BRIMMER test:
The bounds of the standard deviation under constraints.

Jung, L. ([REF]). unsum: Reconstruct raw data from summary statistics with
CLOSURE. R package.

Jung, L., & Heathers, J. ([REF to complete]: CLOSURE algorithm reference.)

Mestdagh, M., Pe, M., Pestman, W., Verdonck, S., Kuppens, P., & Tuerlinckx,
F. (2018). Sidelining the mean: The relative variability index as a generic
mean-corrected variability measure for bounded variables. *Psychological
Methods, 23*(4), 690--707.

Wallrich, L. ([REF]). rsprite2: Identify distributions that match reported
sample parameters (SPRITE). R package.
:::
